    Working with Git Project:
    1. Make sure project is set to D445_Project in RStudio
    2. In RStudio Console:
        getwd() = Verify working directory is set to this project
        setwd() = Can change working directory to this path
        list.files() = Can verify the files are available
    3. In RStudio Terminal:
        cd <destination> = Change directory
        cd - = Go back to previous directory
        ls = Lists contents of current directory
        git init = Initialize git connection
        git status = Check what status needs to be changed before comitting
          - git  add <file>
          - git rm <file>
          - git restore <file>
        git commit -m "" = Commit changes from terminal with message
        git push = Publish local changes
    4. Might need to open GitHub Desktop and verify changes.

```{r}
#############
# Libraries #
#############

library(dplyr)
library(ggplot2)
```

Creating dfs for the price, number of daily bike trips, and the number of daily scooter trips taken over our entire data period.

```{r}
data_dir <- 'FOCO Spin Data'

## https://public.ridereport.com/fortcollins
# df_daily_bike_trips <- data.frame(read.csv(file.path(data_dir,"bikes 2021-07-26--2024-10-17-analyze-trips_by_date.csv")))
# df_hourly_price_rate <- data.frame(read.csv(file.path(data_dir,"spin-prices.csv")))
# date_deploy <- c("2021-07-26", "2022-11-01", "2023-07-01", "2024-07-01")
# scooter_deploy <- c(500, 500, 540, 630)
# df_sccooters_deployed <- data.frame(Date = date_deploy, 
#                                     Scooters = scooter_deploy)

# df_daily_scooter_trips <- data.frame(read.csv(file.path(data_dir,"scooters 2021-07-26--2024-10-17-analyze-trips_by_date.csv")))
# 
# ## data from July 2021 to December 2021
# df_matched_2021_q3 <- data.frame(read.csv(file.path(data_dir, "")), 
#                          stringsAsFactors = FALSE)
# df_matched_2021_q4 <- data.frame(read.csv(file.path(data_dir, "")), 
#                          stringsAsFactors = FALSE)
# df_matched_2022_q1 <- data.frame(read.csv(file.path(data_dir, "")), 
#                          stringsAsFactors = FALSE)
# df_matched_2022_q2 <- data.frame(read.csv(file.path(data_dir, "")), 
#                          stringsAsFactors = FALSE)
# df_matched_2022_q3 <- data.frame(read.csv(file.path(data_dir, "")), 
#                          stringsAsFactors = FALSE)
# df_matched_2022_q4 <- data.frame(read.csv(file.path(data_dir, "")), 
#                          stringsAsFactors = FALSE)
# 
# ## data from January 2023 to June 2024
# df_matched_2023_q1 <- data.frame(read.csv(file.path(data_dir, "")), 
#                          stringsAsFactors = FALSE)
# df_matched_2023_q2 <- data.frame(read.csv(file.path(data_dir, "")), 
#                          stringsAsFactors = FALSE)
# df_matched_2023_q3 <- data.frame(read.csv(file.path(data_dir, "")), 
#                          stringsAsFactors = FALSE)
# df_matched_2023_q4 <- data.frame(read.csv(file.path(data_dir, "")), 
#                          stringsAsFactors = FALSE)
# df_matched_2024_q1 <- data.frame(read.csv(file.path(data_dir, "")), 
#                          stringsAsFactors = FALSE)
# df_matched_2024_q2 <- data.frame(read.csv(file.path(data_dir, "")), 
#                          stringsAsFactors = FALSE)
# 
# ## data from July 2024 to September 2023
# df_matched_2024_q3 <- data.frame(read.csv(file.path(data_dir, "")), 
#                          stringsAsFactors = FALSE)

```

```{r}
# colnames(df_daily_scooter_trips)
df_daily_scooter_trips$Date <- as.Date(df_daily_scooter_trips$Date)
# head(df_daily_scooter_trips)
# str(df_daily_scooter_trips)
daily_date <- df_daily_scooter_trips$Date
daily_trips <- df_daily_scooter_trips$Trips.By.Date...Starts...Spin...Scooters
plot_scooter_rides_daily <- ggplot(data = df_daily_scooter_trips,
                                   mapping = aes(x = daily_date, 
                                                 y = daily_trips)) +
  geom_line() +
  labs(x = "x",
       y = "y",
       title = "title")
plot_scooter_rides_daily
```

```{r}
# head(df_scooter_matched)
# colnames(df_scooter_matched)

df_scooter_no_geo = subset(df_scooter_matched, 
                           select = c("Segment.Name", 
                                      "Percentage.of.Matched.Trips",
                                      "Count.of.Matched.Trips"))
# head(df_scooter_no_geo)
# str(df_scooter_no_geo) ## chr, num, chr
df_scooter_no_geo$Count.of.Matched.Trips <- gsub(",", "", df_scooter_no_geo$Count.of.Matched.Trips)
df_scooter_no_geo$Count.of.Matched.Trips <- gsub("< 100", "50", df_scooter_no_geo$Count.of.Matched.Trips)
df_scooter_no_geo$Count.of.Matched.Trips <- as.numeric(df_scooter_no_geo$Count.of.Matched.Trips)
df_scooter_numeric <- na.omit(df_scooter_no_geo)
# str(df_scooter_numeric)
```
