    Working with Git Project:
    1. Make sure project is set to D445_Project in RStudio
    2. In RStudio Console:
        getwd() = Verify working directory is set to this project
        setwd() = Can change working directory to this path
        list.files() = Can verify the files are available
    3. In RStudio Terminal:
        cd <destination> = Change directory
        cd - = Go back to previous directory
        ls = Lists contents of current directory
        git init = Initialize git connection
        git status = Check what status needs to be changed before comitting
          - git  add <file>
          - git rm <file>
          - git restore <file>
        git commit -m "" = Commit changes from terminal with message
    4. Might need to open GitHub Desktop and verify changes. Then, click "push origin".

```{r}
#############
# Libraries #
#############


```

Creating dfs for the price, number of daily bike trips, and the number of daily scooter trips taken over our entire data period.

```{r}
data_dir <- 'FOCO Spin Data'

df_daily_scooter_trips <- data.frame(read.csv(file.path(data_dir,"scooters 2021-07-26--2024-10-17-analyze-trips_by_date.csv")))
df_daily_bike_trips <- data.frame(read.csv(file.path(data_dir,"bikes 2021-07-26--2024-10-17-analyze-trips_by_date.csv")))
df_hourly_price_rate <- data.frame(read.csv(file.path(data_dir,"spin-prices.csv")))

# head(df_daily_scooter_trips)
# head(df_daily_bike_trips)
# head(df_hourly_price_rate)
```
