This is where I will be exploring the geological start and end locations for the spin cycle datasets provided for our research project. The goals for this portion of the project are:

1.  Use latitude and longitude data to develop a heat map for spin scooter start and end locations.

```{r}
#libraries
library(tidyverse)
library(tidymodels)
library(sf)
library(ggplot2)
library(ggmap)
library(RColorBrewer)
library(patchwork)
library(fields)
library(MASS)
library(dbscan)
library(here)
```

```{r}
set.seed(445)
```

1. Explore the data.

a)  Load the starting and ending latitude and longitude coordinates into respective data-frames. 

```{r}
start_scoot <- read.csv("FOCO Spin Data/fort-collins-curb-events-2021-07-26-2024-10-15-trip_starts.csv")
end_scoot <- read.csv('FOCO Spin Data/fort-collins-curb-events-2021-07-26-2024-10-15-trip_ends.csv')
dpl_scoot <- read.csv('FOCO Spin Data/fort-collins-curb-events-2021-07-26-2024-10-15-deployments.csv')


head(start_scoot)
head(end_scoot)
head(dpl_scoot)
```

```{r}
#observing spatial data range
range(start_scoot$latitude)
range(start_scoot$longitude)

range(end_scoot$latitude)
range(end_scoot$longitude)

end_scoot |> arrange(desc(longitude))
```

b)  can we create a map for our data?

```{r}
#stadiamaps register code
register_stadiamaps('dd6906c8-64ba-46a6-966a-4a3556ae9e34',write = TRUE)
```

```{r}
#set map bounds call stadia map
foco_map_bounds <- c(left = -105.17, bottom = 40.465, right = -104.98, top = 40.7)

foco_map <- get_stadiamap(foco_map_bounds, zoom = 12, maptype = 'stamen_toner_lite')

foco_coord_map <- ggmap(foco_map, extent = 'device', legend = 'none')
foco_coord_map
```
use map bounds to 'clean' the spatial data
```{r}
#clean data
clean_start_scoot <- start_scoot |>
  filter(!is.na(longitude) & !is.na(latitude)) |>
  filter(
    longitude >= foco_map_bounds["left"],
    longitude <= foco_map_bounds["right"],
    latitude >= foco_map_bounds["bottom"],
    latitude <= foco_map_bounds["top"]
  )
clean_end_scoot <- end_scoot |>
  filter(!is.na(longitude) & !is.na(latitude)) |>
  filter(
    longitude >= foco_map_bounds["left"],
    longitude <= foco_map_bounds["right"],
    latitude >= foco_map_bounds["bottom"],
    latitude <= foco_map_bounds["top"]
  )
clean_dpl_scoot <- dpl_scoot |>
  filter(!is.na(longitude) & !is.na(latitude)) |>
  filter(
    longitude >= foco_map_bounds["left"],
    longitude <= foco_map_bounds["right"],
    latitude >= foco_map_bounds["bottom"],
    latitude <= foco_map_bounds["top"]
  )
```



2. can we generate a heatmap for our data?

```{r}
start_map <- foco_coord_map + geom_point(data = clean_start_scoot, aes(x = longitude, y = latitude), alpha = 0.15, size = 0.8, color = 'orange') + theme(panel.grid = element_blank())

start_map

start_dens_heatmap <- foco_coord_map + stat_density2d(data= clean_start_scoot,  aes(x=longitude, y=latitude, fill= after_stat(level)), geom="polygon", contour = TRUE, bins = 500, h = c(0.012, 0.012)) + scale_fill_gradientn(colors = rev(brewer.pal(9, 'PuOr')), name = 'Density') + theme(panel.grid = element_blank())

start_dens_heatmap

start_dens_heatmap + start_map + plot_layout(ncol = 2, guides = 'collect')
```


```{r}
#start heat map-no point map
startmap_refined <- foco_coord_map + stat_density_2d(data = clean_start_scoot, aes(x = longitude, y = latitude, fill = after_stat(level)),
    geom = "polygon",
    contour = TRUE,
    bins = 1000,
    h = c(0.012, 0.012)) + scale_fill_viridis_b(option = 'plasma', name = 'Trip Start\nDensity', guide = guide_colorbar(title.position = "top", barwidth = 1, barheight = 1)) + labs(title = "Fort Collins Scooter Trip Start Hotspots") + theme_minimal() + theme(panel.grid = element_blank(), plot.title = element_text(size = 14, face = "bold", hjust = 0.5), plot.subtitle = element_text(size = 10, hjust = 0.5), legend.position = "right", axis.title = element_blank(), axis.text = element_text(size = 8))

startmap_refined

```
```{r}
end_map <- foco_coord_map + stat_density2d(data= end_scoot,  aes(x=longitude, y=latitude, fill=..level.., alpha=..level..), geom="polygon", contour = TRUE, bins = 15, size = 0.2) + scale_fill_gradientn(colors = rev(brewer.pal(9, 'Spectral'))) +  scale_alpha(range = c(0.2, 0.7)) + geom_point(data = end_scoot, aes(x = longitude, y = latitude), alpha = 0.2, size = 0.5)
```

```{r}
deployment_points <- foco_coord_map + geom_point(data = clean_dpl_scoot, aes(x = longitude, y = latitude), alpha = 0.15, size = 0.8, color = 'orange') + theme(panel.grid = element_blank())
  
deployment_map <- foco_coord_map + stat_density2d(data = clean_dpl_scoot,  aes(x = longitude, y = latitude, fill = after_stat(level)), geom = "polygon", contour = TRUE, bins = 750,h = c(0.012, 0.012)) +
  scale_fill_gradientn(colors = rev(brewer.pal(9, 'PuOr')), name = 'Deployment\nDensity') + 
  theme(panel.grid = element_blank()) + 
  ggtitle("Scooter Deployment Density")

deployment_points + deployment_map + plot_layout(ncol = 2, guides = 'collect')
```

```{r}
#deployment vs. start heatmap
deployment_map + start_dens_heatmap  + 
    plot_layout(ncol = 2, guides = 'collect')

deployment_points + start_map + plot_layout(ncol = 2, guides = 'collect')
```
d) this map provides a basic layout of fort collins. perhaps there is a way to enhance the mapping of our data a little more. (outline key features, form boundaries, draw lines at major roads, etc.)

e) This is a good start to my heatmap but the data feels too separated from each other. There is probably a way to base the density bandwidth on the points in our model. We need our model to base density off of the start points nearest to it instead of the overall density of the data. Right now the density model is saying "I'm in a popular location because I live in Fort Collins." we need each point in the data to say instead "I'm in a popular location because theres a lot of other scooters who started/ended their trip nearby me." 
  
  After doing some research on the topic I'm led to believe that incorporating a KNN model to our data could be very useful here...
```{r}
#set a knn model to compare deployment to start locations to validate that deployments start where they need to.
dpl_knn_model <- nearest_neighbor(neighbors = tune(), weight_func = tune()) |>
  set_engine('kknn') |>
  set_mode('regression')

tuning_grid <- grid_regular(neighbors(range = c(5,25)), weight_func(), levels = c(5,4))

#create grid to count points of data in each grid cell
dpl_dens <- clean_dpl_scoot |>
  mutate(lon_bin = cut_width(longitude, width = 0.005), lat_bin = cut_width(latitude, width = 0.005)) |> 
  group_by(lon_bin, lat_bin) |>
  summarise(n = n(), longitude = mean(longitude), latitude = mean(latitude)) |> 
  mutate(density_score = n/max(n))

#predict density scores using lon + lat
dpl_knn_recipe <- recipe(density_score ~ longitude + latitude, data = dpl_dens)

#c-fold cross validation
cv_folds <- vfold_cv(dpl_dens, v=5)

#knn workflow
dpl_knn_res <- workflow() |>
  add_model(dpl_knn_model) |>
  add_recipe(dpl_knn_recipe) |>
  tune_grid(resamples = cv_folds, grid = tuning_grid)

dpl_best_params <- select_best(dpl_knn_res, metric = 'rmse')

#knn final workflow
dpl_knn_final_fit <- workflow() |>
  add_model(dpl_knn_model) |>
  add_recipe(dpl_knn_recipe) |>
  finalize_workflow(dpl_best_params) |>
  fit(dpl_dens)

#deployment final fit
dpl_knn_final_fit
```
Interpretation:

The model found patterns in how scooters are deployed
Deployments tend to cluster (that's why k=12 works well)
Points influence each other up to 12 neighbors away
Deployment density is somewhat predictable based on location

This model essentially captures Spin's deployment strategy:

Where they think scooters should be
How concentrated their deployments are
The spatial relationship between deployment points

Next steps:
1. build same knn density model for start locations
2. look at the differences


start knn model:
```{r}
#set a knn model to compare deployment to start locations to validate that deployments start where they need to.
start_knn_model <- nearest_neighbor(neighbors = tune(), weight_func = tune()) |>
  set_engine('kknn') |>
  set_mode('regression')

tuning_grid <- grid_regular(neighbors(range = c(5,25)), weight_func(), levels = c(5,4))

#create grid to count points of data in each grid cell
start_dens <- clean_start_scoot |>
  mutate(lon_bin = cut_width(longitude, width = 0.005), lat_bin = cut_width(latitude, width = 0.005)) |> 
  group_by(lon_bin, lat_bin) |>
  summarise(n = n(), longitude = mean(longitude), latitude = mean(latitude)) |> 
  mutate(density_score = n/max(n))

#predict density scores using lon + lat
start_knn_recipe <- recipe(density_score ~ longitude + latitude, data = start_dens)

#c-fold cross validation
cv_folds <- vfold_cv(start_dens, v=5)

#knn workflow
start_knn_res <- workflow() |>
  add_model(start_knn_model) |>
  add_recipe(start_knn_recipe) |>
  tune_grid(resamples = cv_folds, grid = tuning_grid)

start_best_params <- select_best(start_knn_res, metric = 'rmse')

#knn final workflow
start_knn_final_fit <- workflow() |>
  add_model(start_knn_model) |>
  add_recipe(start_knn_recipe) |>
  finalize_workflow(start_best_params) |>
  fit(start_dens)

#deployment final fit
start_knn_final_fit
```

Next Steps:
3. use deployment model to predict what start density 'should' be
4. compare to actual start densities
5. look for areas with big differences
```{r}
#comparing model predictions
grid_pts <- expand.grid(longitude = seq(foco_map_bounds['left'], foco_map_bounds['right'], by = 0.002), latitude = seq(foco_map_bounds['bottom'], foco_map_bounds['top'], by = 0.002))

dpl_pred <- predict(dpl_knn_final_fit, grid_pts)
start_pred <- predict(start_knn_final_fit, grid_pts)

comparison <- grid_pts |>
  mutate(dpl_density = dpl_pred$.pred,
         start_density = start_pred$.pred,
         difference = start_density - dpl_density)

foco_coord_map + geom_tile(data = comparison, aes(x=longitude, y = latitude, fill = difference), alpha = 0.8) + scale_fill_gradient2(low='red',mid='white', high='green', name = 'Deployment vs\nStart Density') + coord_fixed() + theme_minimal()
```
This model is creating a density score comparing the predictions from our deployment and start KNN models. Green areas show zones that could use more scooters deployed and Red areas show areas where deployments may be over-accommodating the area. The right side is showing strong edge effects from the KNN models. Raw data from start locations extend across harmony, mullberry, and e vine up to I-25 but no scooter rides extend east of I-25.
Red means stop, Green means go:
Stop putting so many scooters here (Red)
Go put another scooter here (Green)

The scope of this model stretches across the city of Fort Collins. This can provide a general interpretation of scooter deployment performance across the city, but we will have to zoom in on key mismatching areas to provide a more specific interpretation. Key areas to observe will be spots that show rapid changes in deployment patterns (clusters of red, white, green), and dense red/green areas. Key areas we should look into closer are:
    CSU Campus area
    Old Town area
    Harmony/College intersection
    
We can create smaller maps and confine the model results to these areas to get a clearer picture of where SPIN could add/remove an extra scooter to optimize their deployments.

Start vs. End:
1. Train Start Knn model to test on end locations
2. How well can we predict end locations based on start locations?
```{r}
#Todo: This model doesn't work yet: don't run
st_end_knn_model <- nearest_neighbor(neighbors = tune(), weight_func = tune()) |>
  set_engine('kknn') |>
  set_mode('regression')

tuning_grid <- grid_regular(neighbors(range = c(5,25)), weight_func(), levels = c(5,4))


st_end_knn_recipe <- recipe(density_score_end ~ longitude_start + latitude_start, 
                           data = combined_pts)



#c-fold cross validation
cv_folds <- vfold_cv(combined_pts, v=5)

#knn workflow
st_end_knn_res <- workflow() |>
  add_model(st_end_knn_model) |>
  add_recipe(st_end_knn_recipe) |>
  tune_grid(resamples = cv_folds, grid = tuning_grid)

st_end_best_params <- select_best(st_end_knn_res, metric = 'rmse')

#knn final workflow
st_end_knn_final_fit <- workflow() |>
  add_model(st_end_knn_model) |>
  add_recipe(st_end_knn_recipe) |>
  finalize_workflow(st_end_best_params) |>
  fit(combined_pts)

#deployment final fit
st_end_knn_final_fit

```
Start/End Model Plotting:
```{r}
st_end_pred <- predict(st_end_knn_final_fit, combined_dens)

pred_comparison <- combined_dens |> 
  mutate(pred_end_dens = st_end_pred$.pred,
         actual_end_dens = density_score_end,
         difference = pred_end_dens - actual_end_dens)

foco_coord_map + 
  geom_tile(data = pred_comparison, aes(x = longitude_start, y = latitude_start, fill = difference),) +
  scale_fill_gradient2(low = 'red', mid = 'white', high = 'green', name = 'Predicted vs\nActual End\nDensity') + coord_fixed() + theme_minimal()
```

references: 

building a heatmap using ggmap, stadiamaps, & RColor: <https://axelhodler.medium.com/creating-a-heat-map-from-coordinates-using-r-780db4901075>

nearest neighbor mapping: <https://www.espatial.com/blog/nearest-neighbor-map>

