---
title: "forecasting_weekly_rides"
author: "Grace Combe"
date: "2024-11-03"
output: html_document
---

**GOAL: forecast scooter usage by week over the next year**
```{r}
# setting seed
set.seed(227)
```

```{r}
# load packages
library(tidyverse)
library(tidymodels)
library(randomForest)
library(xgboost)
library(readr)
library(ggplot2)
library(forecast)
library(forecastHybrid)
library(fpp2)
library(TTR)
library(dplyr)
library(lubridate)
library(plotly)
library(rio)
library(tseries)
library(zoo)
library(keras)
library(tensorflow)
```

```{r}
daily_scooters <- read.csv("FOCO Spin Data/scooters 2021-07-26--2024-10-17-analyze-trips_by_date.csv")
daily_scooters <- daily_scooters %>% rename("TripsByDate" = Trips.By.Date...Starts...Spin...Scooters)
```

```{r}
daily_scooters$Date <- ymd(daily_scooters$Date)
daily_scooters$year <- year(daily_scooters$Date)
daily_scooters$WeekNum <- isoweek(ymd(daily_scooters$Date))
```

```{r}
week_data <- daily_scooters %>% group_by(WeekNum, year) %>% summarize(WeeklyRides = sum(TripsByDate)) %>% arrange(year, WeekNum)
```

**Exploratory**
```{r}
#figure out how to connect dots?
ggplot(data = week_data, aes(x = WeekNum, y = WeeklyRides, color = as.factor(year))) +geom_point()
```

**WEEKLY DATA**
```{r}
# make price variable
for (i in 1:nrow(week_data)) {
  week_data$CollegePrice[i] <- ifelse(i <= 71, 0.30, 
                                      ifelse(i <= 154, 0.4, 0.3))
  week_data$NormalPrice[i] <- ifelse(i <= 71, 0.30, 0.40)
}
```

```{r}
# splitting into train/test; test is data from 2024
week_data$Class <- if_else(week_data$year == 2024, 'Test', 'Train')
train <- subset(week_data, Class == 'Train') %>% arrange(year, WeekNum)
test <- subset(week_data, Class == 'Test') %>% arrange(year, WeekNum)
nrow(train)
nrow(test)
```

```{r}
# time series
ts <- ts(train$WeeklyRides, start = c(2021, 30), frequency = 52)
summary(ts)
plot(ts)
```

```{r}
# mean absolute percentage error to measure accuracy
mape <- function(actual, pred){
mape <- mean(abs((actual-pred)/actual))*100
return(mape)
}
```

```{r}
ddata <- decompose(ts, "multiplicative")
plot(ddata)
plot(ddata$trend)
plot(ddata$seasonal)
plot(ddata$random)
plot(ts)
abline(reg = lm(ts ~time(ts)))
boxplot(ts ~ cycle(ts, xlab = "WeekNum", ylab = "WeeklyRides"))
mymodel <- auto.arima(ts)
mymodel
plot.ts(mymodel$residuals)
myforecast <- forecast(mymodel, level = c(95), h = 52)
plot(myforecast)
forecast <-data.frame(ExpApps = as.numeric(myforecast$mean),
                      Lower = as.numeric(myforecast$lower),
                      Upper = as.numeric(myforecast$upper), stringsAsFactors = F)
```

**ARIMA Model**
```{r}
arima_model <- auto.arima(ts, D = 1)
summary(arima_model)
```

```{r}
# mape for test data
fore_arima = forecast::forecast(arima_model, h = 42)
df_arima <- as.data.frame(fore_arima)
test$arima = df_arima$`Point Forecast`
mape(test$WeeklyRides, test$arima[1:42])
```
MAPE 64.95 on test data

**Graphing ARIMA Model**
```{r}
# plot <- ggplot(test, aes(WeekNum)) +
#   geom_line(aes(y = arima, color = "Forecast")) +
#   geom_line(aes(y = WeeklyRides, color = "Actual"))+
#   scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) +
#   ggtitle("Weekly Rides 2024")
# ggplotly(plot)
```
Not really good anywhere

```{r}
# model <- auto.arima(ts, D = 1)
# forecasted_values <- forecast(model, h = 52)
# plot(forecasted_values) # going negative, not good
```

**STLF**
```{r}
stlf_forecast <- stlf(ts, h = 52)
plot(stlf_forecast)
```

```{r}
df_stlf <- as.data.frame(stlf_forecast)
test$stlf = df_stlf$`Point Forecast`[1:42]
mape(test$WeeklyRides, test$stlf)
```
47.12 mape on test (next one is slightly more accurate)

**Graphing STLF**
```{r}
plot <- ggplot(test, aes(WeekNum)) +
  geom_line(aes(y = stlf, color = "Forecast")) +
  geom_line(aes(y = WeeklyRides, color = "Actual"))+
  scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) +
  ggtitle("STLF Weekly Rides 2024")
ggplotly(plot)
```
This gets the spike pretty well, but not the beginning

**Another ARIMA**
```{r}
arima_forecast <- stlf(ts, h = 52, s.window = "periodic", method = "arima")
plot(arima_forecast)
```

```{r}
df_ar <- as.data.frame(arima_forecast)
test$ar = df_ar$`Point Forecast`[1:42]
mape(test$WeeklyRides, test$ar)
```
40.11 mape on test

```{r}
plot <- ggplot(test, aes(WeekNum)) +
  geom_line(aes(y = ar, color = "Forecast")) +
  geom_line(aes(y = WeeklyRides, color = "Actual"))+
  scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) +
  ggtitle("ARIMA Weekly Rides 2024")
ggplotly(plot)
```
Gets the spike pretty well, best so far

**Hybrid Models**
```{r}
library(forecastHybrid)
mod1 <- hybridModel(ts)
mod1_forecast <- forecast(mod1, h = 52)
plot(mod1_forecast)
```

```{r}
df_hybrid <- as.data.frame(mod1_forecast)
test$hybrid = df_hybrid$`Point Forecast`[1:42]
mape(test$WeeklyRides, test$hybrid)
```
33.12 mape on test

```{r}
plot <- ggplot(test, aes(WeekNum)) +
  geom_line(aes(y = hybrid, color = "Forecast")) +
  geom_line(aes(y = WeeklyRides, color = "Actual"))+
  scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) +
  ggtitle("Hybrid Mod1 Weekly Rides 2024")
ggplotly(plot)
```
Doesn't get the spike, but beginning is pretty good

**ARIMA/STLM Hybrid**
```{r}
# ARIMA, STLM
mod_as <- hybridModel(ts, models = "as")
as_fore <- forecast(mod_as, h = 52)
plot(as_fore)
```

```{r}
df_as <- as.data.frame(as_fore)
test$as = df_as$`Point Forecast`[1:42]
mape(test$WeeklyRides, test$as)
```
54.25 mape on test

```{r}
plot <- ggplot(test, aes(WeekNum)) +
  geom_line(aes(y = as, color = "Forecast")) +
  geom_line(aes(y = WeeklyRides, color = "Actual"))+
  scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) + 
  ggtitle("ARIMA + STLM Num Apps 2024")
ggplotly(plot)
```
Gets the spike pretty well, but not so much the beginning

**ARIMA/THETA/NNETAR**
```{r}
# ARIMA, THETA, NNETAR
mod_afn <- hybridModel(ts, models = "afn")
afn_fore <- forecast(mod_afn, h = 52)
plot(afn_fore)
```

```{r}
df_afn <- as.data.frame(afn_fore)
test$afn = df_afn$`Point Forecast`[1:42]
mape(test$WeeklyRides, test$afn)
```
31.88 mape on test

```{r}
plot <- ggplot(test, aes(WeekNum)) +
  geom_line(aes(y = afn, color = "Forecast")) +
  geom_line(aes(y = WeeklyRides, color = "Actual"))+
  scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) +
  ggtitle("ARIMA + THETAM + NNETAR Weekly Rides 2024")
ggplotly(plot)
```
Really good on beginning, but doesn't get spike

**ARIMA/STLM/THETAM**
```{r}
# ARIMA, STLM, THETAM
mod_afs <- hybridModel(ts, models = "afs")
afs_fore <- forecast(mod_afs, h = 52)
plot(afs_fore)
```

```{r}
df_afs <- as.data.frame(afs_fore)
test$afs = df_afs$`Point Forecast`[1:42]
mape(test$WeeklyRides, test$afs)
```
29.69 mape on test data

```{r}
plot <- ggplot(test, aes(WeekNum)) +
  geom_line(aes(y = afs, color = "Forecast")) +
  geom_line(aes(y = WeeklyRides, color = "Actual"))+
  scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) +
  ggtitle("AFS WeeklyRides 2024")
ggplotly(plot)
```
Really good on beginning, but doesn't get spike; BEST SO FAR

**For visualizing price change**
Price dropped to $0.30 for college area on July 1, 2024 (Week 27)
```{r}
plot <- ggplot(test, aes(WeekNum)) +
  geom_line(aes(y = afs, color = "Forecast")) +
  geom_line(aes(y = WeeklyRides, color = "Actual"))+
  scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) +
  ggtitle("AFS WeeklyRides 2024") +
  geom_vline(xintercept = 27, linetype = "dashed", color = "black", size = 0.5)
ggplotly(plot)
```

**Trying to adjust for price did not help**

**one model for pre-price change, one model for after**
```{r}
for (i in 1:nrow(test)){
  test$CombinedPred[i] <- ifelse(test$WeekNum[i] < 27, test$afs[i], test$stlf[i])
}
```

```{r}
mape(test$WeeklyRides, test$CombinedPred)
```
29.04 on test

```{r}
plot <- ggplot(test, aes(WeekNum)) +
  geom_line(aes(y = CombinedPred, color = "Forecast")) +
  geom_line(aes(y = WeeklyRides, color = "Actual"))+
  scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) +
  ggtitle("CombinedPred WeeklyRides 2024") +
  geom_vline(xintercept = 27, linetype = "dashed", color = "black", size = 0.5)
ggplotly(plot)
```
**BEST**

**Non-linear models**
Bagging
```{r}
bagging_spec <- rand_forest(mtry = .cols())  %>% 
  set_engine("randomForest", importance = TRUE) %>% 
  set_mode("regression")

bagging_fit <- bagging_spec %>% 
  fit(WeeklyRides ~ as.factor(WeekNum) + CollegePrice + NormalPrice, data = train)

vip(bagging_fit)

bagging_pred <- predict(bagging_fit, new_data = test)

test$bagging <- bagging_pred$.pred
```

```{r}
mape(test$WeeklyRides, test$bagging)
```
50.37 mape on test

```{r}
plot <- ggplot(test, aes(WeekNum)) +
  geom_line(aes(y = bagging, color = "Forecast")) +
  geom_line(aes(y = WeeklyRides, color = "Actual"))+
  scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) +
  ggtitle("Bagging WeeklyRides 2024")
ggplotly(plot)
```
Did better at getting the spike!, a little too high but right trend everywhere else

Boosting
```{r}
boost_spec <- boost_tree(trees = 5000, tree_depth = 5, learn_rate = 0.01) %>% 
  set_engine("xgboost") %>% 
  set_mode("regression")

boost_fit <- boost_spec %>% 
  fit(WeeklyRides ~ as.factor(WeekNum) + CollegePrice + NormalPrice, data = train)

boost_pred <- predict(boost_fit, new_data = test)

test$boosting <- boost_pred$.pred
```

```{r}
mape(test$WeeklyRides, test$boosting)
```
43.41 mape on test

```{r}
plot <- ggplot(test, aes(WeekNum)) +
  geom_line(aes(y = boosting, color = "Forecast")) +
  geom_line(aes(y = WeeklyRides, color = "Actual"))+
  scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) +
  ggtitle("Boosting WeeklyRides 2024")
ggplotly(plot)
```
captures spike but needs to be better in beginning

Random Forest--Not good
```{r}
rf_spec <- rand_forest(mtry = sqrt(.cols()))  %>%
  set_engine("randomForest", importance = TRUE) %>%
  set_mode("regression")

rf_fit <- rf_spec %>%
  fit(WeeklyRides ~ as.factor(WeekNum) + CollegePrice + NormalPrice, data = train)

rf_pred <- predict(rf_fit, new_data = test)

test$rf <- rf_pred$.pred
```

```{r}
mape(test$WeeklyRides, test$rf)
```
74.81 mape on test

```{r}
plot <- ggplot(test, aes(WeekNum)) +
  geom_line(aes(y = rf, color = "Forecast")) +
  geom_line(aes(y = WeeklyRides, color = "Actual"))+
  scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) +
  ggtitle("RF WeeklyRides 2024")
ggplotly(plot)
```


-------------------None of this is good, ignore-----------------------------
**Trying to improve best model from before (ARIMA/STLM/THETAM) with boosting/bagging**
**Neither of these worked well: both way too low**
```{r}
# # Get residuals from the hybrid model
# residuals_afs <- residuals(mod_afs)
# 
# # Create lag features from the residuals
# residuals_df <- data.frame(
#   residuals = as.vector(residuals_afs),
#   lag1 = as.numeric(lag(as.vector(residuals_afs), 1)),
#   lag2 = as.numeric(lag(as.vector(residuals_afs), 2))
# )
# residuals_df <- na.omit(residuals_df)  # Remove NAs created by lagging
# 
# # Define target and predictors
# target_residual <- residuals_df$residuals
# predictors_residual <- residuals_df[, -1]
# 
# # Train an XGBoost model on the residuals
# dtrain <- xgb.DMatrix(data = as.matrix(predictors_residual), label = target_residual)
# params <- list(objective = "reg:squarederror", max_depth = 5, eta = 0.1)
# xgb_model <- xgboost(params = params, data = dtrain, nrounds = 100)
# 
# # Predict the residuals for the forecast period
# lagged_residuals <- residuals_afs[(length(residuals_afs) - 1):(length(residuals_afs) - 2)]
# predictions_residuals <- predict(xgb_model, as.matrix(lagged_residuals))
# 
# # Add predicted residuals to the hybrid forecast
# improved_forecast <- as.data.frame(afs_fore$mean + predictions_residuals)
# 
# # Plot the improved forecast
# plot(improved_forecast, main = "Improved Forecast with XGBoost on Residuals")
```

```{r}
# test$other_boost <- improved_forecast$x[1:42]
```

```{r}
# mape(test$WeeklyRides, test$other_boost)
```
62.001 mape on test

```{r}
# plot <- ggplot(test, aes(WeekNum)) +
#   geom_line(aes(y = other_boost, color = "Forecast")) +
#   geom_line(aes(y = WeeklyRides, color = "Actual"))+
#   scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) +
#   ggtitle("AFS with Boosting WeeklyRides 2024")
# ggplotly(plot)
```

This one did not work well
```{r}
# # Fit additional models (ARIMA, STLM, Theta)
# mod_arima <- auto.arima(ts)
# mod_stlm <- stlm(ts)
# mod_theta <- thetaf(ts, h = 52)
# 
# # Forecast with each model
# forecast_arima <- forecast(mod_arima, h = 52)$mean
# forecast_stlm <- forecast(mod_stlm, h = 52)$mean
# forecast_theta <- mod_theta$mean
# 
# # Combine the predictions from different models
# combined_predictions <- data.frame(
#   hybrid = afs_fore$mean,
#   arima = forecast_arima,
#   stlm = forecast_stlm,
#   theta = forecast_theta
# )
# 
# # Train an XGBoost model to combine the forecasts
# dtrain <- xgb.DMatrix(data = as.matrix(combined_predictions), label = afs_fore$mean)
# params <- list(objective = "reg:squarederror", max_depth = 5, eta = 0.1)
# xgb_model <- xgboost(params = params, data = dtrain, nrounds = 100)
# 
# # Make final forecast using the ensemble model
# final_predictions <- as.data.frame(predict(xgb_model, as.matrix(combined_predictions)))
```

```{r}
# test$afs_boost <- final_predictions$`predict(xgb_model, as.matrix(combined_predictions))`[1:42]
```

```{r}
# mape(test$WeeklyRides, test$afs_boost)
```
67.471 mape on test

```{r}
# plot <- ggplot(test, aes(WeekNum)) +
#   geom_line(aes(y = afs_boost, color = "Forecast")) +
#   geom_line(aes(y = WeeklyRides, color = "Actual"))+
#   scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) +
#   ggtitle("AFS with Boosting WeeklyRides 2024")
# ggplotly(plot)
```
Way too low

**Auto ARIMA**
Exploratory
```{r}
# train_ts <- ts(train$WeeklyRides)
# train$clean_count <- tsclean(train_ts) # cleans missing data, really bad outliers, etc
```

```{r}
# plot cleaned data; not as many outliers
#ggplot() + geom_line(data = train, aes(x = WeekEndingDate, y = clean_count)) + ylab('Cleaned Count Apps')
# plot uncleaned data
#ggplot() + geom_line(data = train, aes(x = WeekEndingDate, y = numApps)) + ylab('Uncleaned Count Apps')
```

```{r}
# moving averages
# train$count_ma = ma(train$clean_count, order = 7)
# train$count_ma30 = ma(train$clean_count, order = 30)
# train <- replace(train, TRUE, lapply(train, na.aggregate)) #fills in NAs with column mean
```

```{r}
# plot with NAs removed
# ggplot() +
#   geom_line(data = train, aes(x = WeekEndingDate, y = clean_count, colour = "Counts")) +
#   geom_line(data = train, aes(x = WeekEndingDate, y = count_ma, colour = "Weekly Moving Average")) +
#   geom_line(data = train, aes(x = WeekEndingDate, y = count_ma30, colour = "Monthly Moving Average")) +
#   ylab('Cleaned Count Apps')
```

Decomposition
```{r}
# count_ma <- ts(na.omit(train$clean_count), frequency = 52)
# decomp <- stl(count_ma, "periodic")
# deseasonal_count <- seasadj(decomp) #used in ARIMA model later
# plot(decomp)
# clear seasonality
```

```{r}
# test for stationarity
# adf.test(count_ma, alternative = "stationary")
# p value of less than 0.05 indicates stationary data set; 0.07
```

```{r}
# auto correlation and seasonal differencing
# ACF plots display correlation between a series and its lags; don't want to go outside dotted lines, if so = lags
# Acf(count_ma, main = '') # outside lines
# PACF plots display correlation between a series and its lags that explained by previous lags
# Pacf(count_ma, main = '') # only one outside lines
```

```{r}
# use d = 1 if in bounds, d = 2 if out of bounds (try to use 1 unless you can't)
# count_d2 <- diff(deseasonal_count, differences = 2)
# plot(count_d2)
```

```{r}
# retest with adf
# adf.test(count_d2, alternative = "stationary") # yay less than 0.05 now; we want to be as stationary as possible
```

```{r}
# look for spikes at specific lag points of the differenced series (using count_d2 from above)
# Acf(count_d2, main = 'ACF for Differenced Series') # biggest lag at 1
# Pacf(count_d2, main = 'PACF for Differenced Series') # biggest lag at 1 again
```

Fitting ARIMA model
```{r}
# get auto fit p, d, q models
# auto.arima(deseasonal_count, seasonal = FALSE)
```

```{r}
# evaluate and iterate - does the model make sense?
# fit <- auto.arima(deseasonal_count, seasonal = FALSE)
# tsdisplay(residuals(fit), lag.max = 52, main = '(2,1,1) Model Residuals') # looks like biggest lag at 52
```

```{r}
# adjust for diff = 2, lag at 52
# fit2 <- arima(deseasonal_count, order = c(0,2,52))
# tsdisplay(residuals(fit2), lag.max = 52, main = 'Seasonal Model Residuals')
```

```{r}
# train$deseasonal_count <- deseasonal_count
```

```{r}
# testing
# fit2_forecast <- data.frame(forecast(fit2, h = 42))
# test$fit2 <- fit2_forecast$Point.Forecast
```

```{r}
# mape(test$WeeklyRides, test$fit2)
```
113.06 mape on test

```{r}
# plot <- ggplot(test, aes(WeekNum)) +
#   geom_line(aes(y = fit2, color = "Forecast")) +
#   geom_line(aes(y = WeeklyRides, color = "Actual"))+
#   scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) +
#   ggtitle("Fit2 ARIMA Weekly Rides 2024")
# ggplotly(plot)
```

```{r}
# bring back seasonality?
# seas_fit <- auto.arima(deseasonal_count, seasonal = TRUE)
# seas_forecast <- forecast(seas_fit, h = 42)
# plot(seas_forecast) # straight line is bad
```

Further testing
```{r}
# adjust for diff = 2, lag at 52
# fit3 <- arima(deseasonal_count, order = c(1,2,52))
# tsdisplay(residuals(fit3), lag.max = 52, main = 'Seasonal Model Residuals') # lags look the same
```

```{r}
# testing
# fit3_forecast <- data.frame(forecast(fit3, h = 42))
# test$fit3 <- fit3_forecast$Point.Forecast
```

```{r}
# mape(test$WeeklyRides, test$fit3)
```
112.29 mape on test

```{r}
# plot <- ggplot(test, aes(WeekNum)) +
#   geom_line(aes(y = fit3, color = "Forecast")) +
#   geom_line(aes(y = WeeklyRides, color = "Actual"))+
#   scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) +
#   ggtitle("Fit3 ARIMA Weekly Rides 2024")
# ggplotly(plot)
```

**REFITTING MODELS WITH CLEANED DATA**
```{r}
# ts_clean <- ts(train$clean_count, start = c(2021, 1), frequency = 52)
```

```{r}
# ddata <- decompose(ts_clean, "multiplicative")
# plot(ddata)
# plot(ddata$trend)
# plot(ddata$seasonal)
# plot(ddata$random)
# plot(ts_clean)
# abline(reg = lm(ts_clean ~time(ts_clean)))
# boxplot(ts_clean ~ cycle(ts_clean, xlab = "WeekOfYear", ylab = "numApps"))
# mymodel <- auto.arima(ts_clean)
# mymodel
# plot.ts(mymodel$residuals)
# myforecast <- forecast(mymodel, level = c(95), h = 52)
# plot(myforecast)
# forecast <-data.frame(ExpApps = as.numeric(myforecast$mean),
#                       Lower = as.numeric(myforecast$lower),
#                       Upper = as.numeric(myforecast$upper), stringsAsFactors = F)
```

**ARIMA Model**
```{r}
# arima_model <- auto.arima(ts_clean, D = 1)
# summary(arima_model)
```

```{r}
# mape for test data
# fore_arima = forecast::forecast(arima_model, h = 42)
# df_arima <- as.data.frame(fore_arima)
# test$arima = df_arima$`Point Forecast`
# mape(test$WeeklyRides, test$arima[1:42])
```
66.40 mape on test

**Graphing ARIMA Model**
```{r}
# plot <- ggplot(test, aes(WeekNum)) +
#   geom_line(aes(y = arima, color = "Forecast")) +
#   geom_line(aes(y = WeeklyRides, color = "Actual"))+
#   scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) +
#   ggtitle("Weekly Rides 2024")
# ggplotly(plot)
```

**Different ARIMA**
```{r}
# model <- auto.arima(ts_clean, D = 1)
# forecasted_values <- forecast(model, h = 52)
# plot(forecasted_values)
```

**STLF on clean**
```{r}
# stlf_forecast <- stlf(ts_clean, h = 52)
# plot(stlf_forecast)
```

```{r}
# df_stlf <- as.data.frame(stlf_forecast)
# test$stlf_clean = df_stlf$`Point Forecast`[1:42]
# mape(test$WeeklyRides, test$stlf_clean)
```
46.39 mape on test

**Graphing STLF**
```{r}
# plot <- ggplot(test, aes(WeekNum)) +
#   geom_line(aes(y = stlf_clean, color = "Forecast")) +
#   geom_line(aes(y = WeeklyRides, color = "Actual"))+
#   scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) +
#   ggtitle("STLF Weekly Rides 2024")
# ggplotly(plot)
```

```{r}
# arima_forecast <- stlf(ts_clean, h = 52, s.window = "periodic", method = "arima")
# plot(arima_forecast)
```

```{r}
# df_ar <- as.data.frame(arima_forecast)
# test$ar_clean = df_ar$`Point Forecast`[1:42]
# mape(test$WeeklyRides, test$ar_clean)
```
42.31 mape on test

```{r}
# plot <- ggplot(test, aes(WeekNum)) +
#   geom_line(aes(y = ar_clean, color = "Forecast")) +
#   geom_line(aes(y = WeeklyRides, color = "Actual"))+
#   scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) +
#   ggtitle("ARIMA Weekly Rides 2024")
# ggplotly(plot)
```

**Hybrid Models**
```{r}
# library(forecastHybrid)
# mod1 <- hybridModel(ts_clean)
# mod1_forecast <- forecast(mod1, h = 52)
# plot(mod1_forecast)
```

```{r}
# df_hybrid2 <- as.data.frame(mod1_forecast)
# test$hybrid2 = df_hybrid2$`Point Forecast`[1:42]
# mape(test$WeeklyRides, test$hybrid2)
```
33.58 mape on test

```{r}
# plot <- ggplot(test, aes(WeekNum)) +
#   geom_line(aes(y = hybrid2, color = "Forecast")) +
#   geom_line(aes(y = WeeklyRides, color = "Actual"))+
#   scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) +
#   ggtitle("Hybrid2 Weekly Rides 2024")
# ggplotly(plot)
```

**ARIMA/STLM**
```{r}
# ARIMA, STLM
# mod_as_clean <- hybridModel(ts_clean, models = "as") # no ets bc h >24
# as_fore_clean <- forecast(mod_as_clean, h = 52)
# plot(as_fore_clean)
```

```{r}
# df_as_clean <- as.data.frame(as_fore_clean)
# test$as_clean = df_as$`Point Forecast`[1:42]
# mape(test$WeeklyRides, test$as_clean)
```
54.25 mape on test

```{r}
# plot <- ggplot(test, aes(WeekNum)) +
#   geom_line(aes(y = as_clean, color = "Forecast")) +
#   geom_line(aes(y = WeeklyRides, color = "Actual"))+
#   scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) +
#   ggtitle("ARIMA + STLM Weekly Rides 2024")
# ggplotly(plot)
```
---------------------------------------------------------------
