    Working with Git Project:
    1. Make sure project is set to D445_Project in RStudio
    2. In RStudio Console:
        getwd() = Verify working directory is set to this project
        setwd() = Can change working directory to this path
        list.files() = Can verify the files are available
    3. In RStudio Terminal:
        cd <destination> = Change directory
        cd - = Go back to previous directory
        ls = Lists contents of current directory
        git init = Initialize git connection
        git status = Check what status needs to be changed before comitting
          - git  add <file>
          - git rm <file>
          - git restore <file>
        git commit -m "" = Commit changes from terminal with message
        git push = Publish local changes
    4. Might need to open GitHub Desktop and verify changes.

```{r, warning=F, message=F}
#############
# Libraries #
#############

library(dplyr)
library(forecast)
library(ggplot2)
library(htmltools)
library(leaflet)
library(lubridate)
library(mapview)
library(prophet) 
library(sf)
library(tidyverse)
library(tmap)
library(webshot)
```

Creating dfs for the price, number of daily bike trips, and the number of daily scooter trips taken over our entire data period.

```{r}
data_dir <- 'FOCO Spin Data'
d2 <- 'FOCO Spin Data/Quartely Matched Trip Files'

## https://public.ridereport.com/fortcollins
# df_daily_bike_trips <- data.frame(read.csv(file.path(data_dir,"bikes 2021-07-26--2024-10-17-analyze-trips_by_date.csv")))
# df_hourly_price_rate <- data.frame(read.csv(file.path(data_dir,"spin-prices.csv")))
# date_deploy <- c("2021-07-26", "2022-11-01", "2023-07-01", "2024-07-01")
# scooter_deploy <- c(500, 500, 540, 630)
# df_sccooters_deployed <- data.frame(Date = date_deploy, 
#                                     Scooters = scooter_deploy)

df_daily_scooter_trips <- data.frame(read.csv(file.path(data_dir,"scooters 2021-07-26--2024-10-17-analyze-trips_by_date.csv")))

## data from July 2021 to December 2021
df_matched_2021_q3 <- data.frame(read.csv(file.path(d2, "fortcollins-routes-data-for-scooter-in-2021-Q3.csv")),
                         stringsAsFactors = FALSE)
df_matched_2021_q4 <- data.frame(read.csv(file.path(d2, "fortcollins-routes-data-for-scooter-in-2021-Q4.csv")),
                         stringsAsFactors = FALSE)
df_matched_2022_q1 <- data.frame(read.csv(file.path(d2, "fortcollins-routes-data-for-scooter-in-2022-Q1.csv")),
                         stringsAsFactors = FALSE)
df_matched_2022_q2 <- data.frame(read.csv(file.path(d2, "fortcollins-routes-data-for-scooter-in-2022-Q2.csv")),
                         stringsAsFactors = FALSE)
df_matched_2022_q3 <- data.frame(read.csv(file.path(d2, "fortcollins-routes-data-for-scooter-in-2022-Q3.csv")),
                         stringsAsFactors = FALSE)
df_matched_2022_q4 <- data.frame(read.csv(file.path(d2, "fortcollins-routes-data-for-scooter-in-2022-Q4.csv")),
                         stringsAsFactors = FALSE)

## data from January 2023 to June 2024
df_matched_2023_q1 <- data.frame(read.csv(file.path(d2, "fortcollins-routes-data-for-scooter-in-2023-Q1.csv")),
                         stringsAsFactors = FALSE)
df_matched_2023_q2 <- data.frame(read.csv(file.path(d2, "fortcollins-routes-data-for-scooter-in-2023-Q2.csv")),
                         stringsAsFactors = FALSE)
df_matched_2023_q3 <- data.frame(read.csv(file.path(d2, "fortcollins-routes-data-for-scooter-in-2023-Q3.csv")),
                         stringsAsFactors = FALSE)
df_matched_2023_q4 <- data.frame(read.csv(file.path(d2, "fortcollins-routes-data-for-scooter-in-2023-Q4.csv")),
                         stringsAsFactors = FALSE)
df_matched_2024_q1 <- data.frame(read.csv(file.path(d2, "fortcollins-routes-data-for-scooter-in-2024-Q1.csv")),
                         stringsAsFactors = FALSE)
df_matched_2024_q2 <- data.frame(read.csv(file.path(d2, "fortcollins-routes-data-for-scooter-in-2024-Q2.csv")),
                         stringsAsFactors = FALSE)

## data from July 2024 to September 2023
df_matched_2024_q3 <- data.frame(read.csv(file.path(d2, "fortcollins-routes-data-for-scooter-in-2024-Q3.csv")),
                         stringsAsFactors = FALSE)

```

```{r, fig.width=14, fig.height=6}
price_change_1 <- as.Date("2021-07-26")
price_change_2 <- as.Date("2022-12-01")
price_change_3 <- as.Date("2024-07-01")
df_daily_scooter_trips$Date <- as.Date(df_daily_scooter_trips$Date)
daily_date <- df_daily_scooter_trips$Date
daily_trips <- df_daily_scooter_trips$Trips.By.Date...Starts...Spin...Scooters
plot_scooter_rides_daily <- ggplot(data = df_daily_scooter_trips,
                                   mapping = aes(x = daily_date, 
                                                 y = daily_trips)) +
  geom_line(alpha = 0.5) +
  theme_minimal() +
  scale_x_continuous(breaks = pretty(daily_date, n = 8)) +
  geom_vline(xintercept = price_change_1, color = "blue") +
  geom_vline(xintercept = price_change_2, color = "blue") +
  geom_vline(xintercept = price_change_3, color = "blue") +
  annotate("text", x = as.Date("2021-07-26"), 
           y = max(daily_trips, na.rm = TRUE),
           label = "Initial Price\n(0.30 $/min)\nJuly 26, 2021", 
           angle = 0, hjust = -0.1, size = 2) +
  annotate("text", x = as.Date("2022-12-01"), 
           y = max(daily_trips, na.rm = TRUE),
           label = "First Price\nChange (0.40 $/min)\nDec 1, 2022", 
           angle = 0, hjust = -0.1, size = 2) +
  annotate("text", x = as.Date("2024-07-01"), 
           y = max(daily_trips, na.rm = TRUE),
           label = "Second Price\nChange (0.40 $/min,\n0.30 $/min on campus)\nJuly 1, 2024", 
           angle = 0, hjust = -0.1, size = 2) +
  labs(x = "Date",
       y = "Rides",
       title = "Number of Rides Taken on Spin Scooters")
plot_scooter_rides_daily
```

```{r}
df_p1 <- rbind(df_matched_2021_q3, df_matched_2021_q4,
               df_matched_2022_q1, df_matched_2022_q2,
               df_matched_2022_q3, df_matched_2022_q4)

df_p2 <- rbind(df_matched_2023_q1, df_matched_2023_q2,
               df_matched_2023_q3, df_matched_2023_q4,
               df_matched_2024_q1, df_matched_2024_q2)

df_p3 <- df_matched_2024_q3

data_list <- list(df_p1, df_p2, df_p3)

for (i in seq_along(data_list)) {
  df <- data_list[[i]]
  df$Count.of.Matched.Trips <- gsub(",", "", df$Count.of.Matched.Trips)
  df$Count.of.Matched.Trips <- gsub("< 100", "50",
                                    df$Count.of.Matched.Trips)
  df$Count.of.Matched.Trips <- as.numeric(df$Count.of.Matched.Trips)
  df <- na.omit(df)

  df <- df %>%
    group_by(Segment.Name) %>%
    summarize(
      Total_Matched_Trips = sum(Count.of.Matched.Trips),
      Weighted_Average_Percentage = sum(Count.of.Matched.Trips *
                                          Percentage.of.Matched.Trips) /
        sum(Count.of.Matched.Trips)
      )
  df$Price_Period <- i
  df <- rename(df, Segment_Name = Segment.Name)
  data_list[[i]] <- df
}

df_p1 <- data_list[[1]]
df_p2 <- data_list[[2]]
df_p3 <- data_list[[3]]

# head(df_p1)

foco <- st_read("~/GitHub/D445_Project/FOCO Spin Data/Street_Centerlines")

# head(foco)

foco$STRNAME <- str_to_upper(str_trim(foco$STRNAME))
foco <- foco %>%
  arrange(STRNAME)
foco$STRNAME <- gsub(" ST", " STREET", foco$STRNAME)
foco$STRNAME <- gsub(" AVE", " AVENUE", foco$STRNAME)
foco$STRNAME <- gsub(" BLVD", " BOULEVARD", foco$STRNAME)
foco$STRNAME <- gsub(" CT", " COURT", foco$STRNAME)
foco$STRNAME <- gsub(" DR", " DRIVE", foco$STRNAME)
foco$STRNAME <- gsub(" CIR", " CIRCLE", foco$STRNAME)
foco$STRNAME <- gsub(" LN", " LANE", foco$STRNAME)
foco$STRNAME <- gsub(" MT", " MOUNT", foco$STRNAME)
foco$STRNAME <- gsub(" PKWY", " PARKWAY", foco$STRNAME)
foco$STRNAME <- gsub(" RD", " ROAD", foco$STRNAME)
foco$STRNAME <- gsub("N ", "NORTH ", foco$STRNAME)
foco$STRNAME <- gsub("S ", "SOUTH ", foco$STRNAME)
foco$STRNAME <- gsub("E ", "EAST ", foco$STRNAME)
foco$STRNAME <- gsub("W ", "WEST ", foco$STRNAME)
df_p1$Segment_Name <- str_to_upper(str_trim(df_p1$Segment_Name))
df_p2$Segment_Name <- str_to_upper(str_trim(df_p2$Segment_Name))
df_p3$Segment_Name <- str_to_upper(str_trim(df_p3$Segment_Name))

join_data <- foco %>%
  left_join(df_p1, by = c("STRNAME" = "Segment_Name"))

# sum(!is.na(join_data$Total_Matched_Trips))

join_data <- join_data %>%
  filter(!is.na(Total_Matched_Trips))
join_data$Total_Matched_Trips <- as.numeric(join_data$Total_Matched_Trips)

# head(join_data)

join_data2 <- foco %>%
  left_join(df_p2, by = c("STRNAME" = "Segment_Name"))
join_data2 <- join_data2 %>%
  filter(!is.na(Total_Matched_Trips))
join_data2$Total_Matched_Trips <- as.numeric(join_data2$Total_Matched_Trips)

join_data3 <- foco %>%
  left_join(df_p3, by = c("STRNAME" = "Segment_Name"))
join_data3 <- join_data3 %>%
  filter(!is.na(Total_Matched_Trips))
join_data3$Total_Matched_Trips <- as.numeric(join_data3$Total_Matched_Trips)
```
```{r}

join_data <- st_transform(join_data, crs = 4326)

color_pal <- colorNumeric(
  palette = "viridis",
  domain = range(join_data$Weighted_Average_Percentage, na.rm = TRUE)
)

a <- leaflet(join_data) %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolylines(color = ~color_pal(Weighted_Average_Percentage),
               weight = 2,
               opacity = 0.9,
               popup = ~paste("Street Name:", 
                              STRNAME, 
                              "<br>Total Matched Trips:",
                              Total_Matched_Trips,
                              "<br>Weighted Average Percentage:",
                              Weighted_Average_Percentage)
               ) %>%
  addLegend(position = "bottomright", 
            pal = color_pal, 
            values = range(join_data$Weighted_Average_Percentage), 
            title = HTML("First Price Period<br>Weighted Percentage of<br>Matched Trips"),
            labFormat = labelFormat(suffix = "%")
            )
a
```
```{r}
join_data2 <- st_transform(join_data2, crs = 4326)

b <- leaflet(join_data2) %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolylines(color = ~color_pal(Weighted_Average_Percentage),
               weight = 2,
               opacity = 0.9,
               popup = ~paste("Street Name:", 
                              STRNAME, 
                              "<br>Total Matched Trips:",
                              Total_Matched_Trips,
                              "<br>Weighted Average Percentage:",
                              Weighted_Average_Percentage)
               ) %>%
  addLegend(position = "bottomright", 
            pal = color_pal, 
            values = range(join_data$Weighted_Average_Percentage),
            title = HTML("Second Price Period<br>Weighted Percentage of<br>Matched Trips"),
            labFormat = labelFormat(suffix = "%")
            )
b
```
```{r}
join_data3 <- st_transform(join_data3, crs = 4326)

c <- leaflet(join_data3) %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolylines(color = ~color_pal(Weighted_Average_Percentage),
               weight = 2,
               opacity = 0.9,
               popup = ~paste("Street Name:", 
                              STRNAME, 
                              "<br>Total Matched Trips:",
                              Total_Matched_Trips,
                              "<br>Weighted Average Percentage:",
                              Weighted_Average_Percentage)
               ) %>%
  addLegend(position = "bottomright", 
            pal = color_pal, 
            values = range(join_data$Weighted_Average_Percentage),
            title = HTML("Third Price Period<br>Weighted Percentage of<br>Matched Trips"),
            labFormat = labelFormat(suffix = "%")
            )
c
```

```{r}
# df_matched_trips <- rbind(df_p1, df_p2, df_p3)
# 
# df_matched_trips <- df_matched_trips %>%
#     mutate(
#       Normalized_Weighted_Avg_Percentage = (Weighted_Average_Percentage - min(Weighted_Average_Percentage)) / (max(Weighted_Average_Percentage) - min(Weighted_Average_Percentage))
#       )
# 
# # head(df_matched_trips)
```
```{r}
# df_matched_2021_q3$Quarter<- "2021 Q3"
# df_matched_2021_q3$Price_Period <- "1"
# df_matched_2021_q4$Quarter<- "2021 Q4"
# df_matched_2021_q4$Price_Period <- "1"
# df_matched_2022_q1$Quarter<- "2022 Q1"
# df_matched_2022_q1$Price_Period <- "1"
# df_matched_2022_q2$Quarter<- "2022 Q2"
# df_matched_2022_q2$Price_Period <- "1"
# df_matched_2022_q3$Quarter<- "2022 Q3"
# df_matched_2022_q3$Price_Period <- "1"
# df_matched_2022_q4$Quarter<- "2022 Q4"
# df_matched_2022_q4$Price_Period <- "1"
# 
# df_matched_2023_q1$Quarter<- "2023 Q1"
# df_matched_2023_q1$Price_Period <- "2"
# df_matched_2023_q2$Quarter<- "2023 Q2"
# df_matched_2023_q2$Price_Period <- "2"
# df_matched_2023_q3$Quarter<- "2023 Q3"
# df_matched_2023_q3$Price_Period <- "2"
# df_matched_2023_q4$Quarter<- "2023 Q4"
# df_matched_2023_q4$Price_Period <- "2"
# df_matched_2024_q1$Quarter<- "2024 Q1"
# df_matched_2024_q1$Price_Period <- "2"
# df_matched_2024_q2$Quarter<- "2024 Q1"
# df_matched_2024_q2$Price_Period <- "2"
# 
# df_matched_2024_q3$Quarter<- "2024 Q3"
# df_matched_2024_q3$Price_Period <- "3"
# 
# data_list2 <- list(df_matched_2021_q3, df_matched_2021_q4,
#                df_matched_2022_q1, df_matched_2022_q2,
#                df_matched_2022_q3, df_matched_2022_q4,
#                df_matched_2023_q1, df_matched_2023_q2,
#                df_matched_2023_q3, df_matched_2023_q4,
#                df_matched_2024_q1, df_matched_2024_q2,
#                df_matched_2024_q3)
# 
# for (i in seq_along(data_list2)) {
#   df <- data_list2[[i]]
#   df <- subset(df, select = c(Segment.Name, Percentage.of.Matched.Trips,
#                               Count.of.Matched.Trips, Quarter,
#                               Price_Period))
#   df$Count.of.Matched.Trips <- gsub(",", "", df$Count.of.Matched.Trips)
#   df$Count.of.Matched.Trips <- gsub("< 100", "50", 
#                                     df$Count.of.Matched.Trips)
#   df$Count.of.Matched.Trips <- as.numeric(df$Count.of.Matched.Trips)
#   df <- na.omit(df)
#   
#   df <- df %>%
#     group_by(Segment.Name, Quarter, Price_Period) %>%
#     summarize(
#       Total_Matched_Trips = sum(Count.of.Matched.Trips),
#       Weighted_Average_Percentage = sum(Count.of.Matched.Trips *
#                                           Percentage.of.Matched.Trips) /
#         sum(Count.of.Matched.Trips)
#       ) %>% 
#     ungroup()
#   df <- rename(df, Segment_Name = Segment.Name)
#   data_list2[[i]] <- df
# }
# a <- data_list2[[1]]
# b <- data_list2[[2]]
# c <- data_list2[[3]]
# d <- data_list2[[4]]
# e <- data_list2[[5]]
# f <- data_list2[[6]]
# g <- data_list2[[7]]
# h <- data_list2[[8]]
# i <- data_list2[[9]]
# j <- data_list2[[10]]
# k <- data_list2[[11]]
# l <- data_list2[[12]]
# m <- data_list2[[13]]
# 
# test_df <- rbind(a, b, c, d, e, f, g, h, i, j, k, l, m)
# test_df <- test_df %>% 
#   filter(!is.na(Segment_Name) & Segment_Name != "")
# head(test_df)
```
```{r}
# splice_plot <- ggplot(data = df_matched_trips, aes(x = as.numeric(as.factor(Price_Period)), y = Normalized_Weighted_Avg_Percentage)) +
#   geom_point() +
#   geom_smooth(method = "lm", formula = y ~ ns(x, df = 3), se = FALSE) +  
#   # geom_vline(xintercept = c(2, 3), linetype = "dashed", color = "red") +  
#   # annotate("text", x = 2, y = max(combined_data$Normalized_Weighted_Avg_Percentage) * 0.95, label = "Price Change 1", color = "red", angle = 90, vjust = 1) +  
#   # annotate("text", x = 3, y = max(combined_data$Normalized_Weighted_Avg_Percentage) * 0.95, label = "Price Change 2", color = "red", angle = 90, vjust = 1) +
#   theme_light() +
#   labs(
#     title = "Returning Users with Price Changes Highlighted",
#     x = "Quarter",
#     y = "Normalized Percentage of Returning Users",
#     color = "Street Segment"
#   )
# plot(splice_plot)
```

```{r}
df_daily_scooter_trips$Date <- as.Date(df_daily_scooter_trips$Date)
df_daily_scooter_trips$price = NA

price_hehe <- data.frame(
  start_date = as.Date(c('2021-07-26', '2022-12-01', '2024-07-01')),  # Price change start dates
  end_date = as.Date(c('2022-11-30', '2024-06-30', '2024-10-17')),    # Price change end dates
  price = c(0.30, 0.40, 0.35)  # Prices during each time period
)

df_daily_scooter_trips <- df_daily_scooter_trips %>%
  mutate(price = case_when(
    Date >= price_hehe$start_date[1] & Date <= price_hehe$end_date[1] ~
      price_hehe$price[1],
    Date >= price_hehe$start_date[2] & Date <= price_hehe$end_date[2] ~
      price_hehe$price[2],
    Date >= price_hehe$start_date[3] & Date <= price_hehe$end_date[3] ~
      price_hehe$price[3],
    TRUE ~ NA_real_)
  )

# head(df_daily_scooter_trips)
```
```{r}
actual_rides_data <- df_daily_scooter_trips %>%
  filter(Date >= as.Date("2021-07-01")) %>%
  select(Date, actual_rides = Trips.By.Date...Starts...Spin...Scooters)

future_dates <- data.frame(Date = seq(as.Date("2024-07-01"), 
                                      as.Date("2024-10-17"), 
                                      by = "day"))
future_dates$price <- 0.40

## Linear Model
training_data <- df_daily_scooter_trips %>%
  filter(Date < as.Date("2024-07-01")) %>%
  mutate(Trip_Occurred = ifelse(Trips.By.Date...Starts...Spin...Scooters > 0, 1, 0))
model <- lm(Trips.By.Date...Starts...Spin...Scooters ~ Date + 
              price, data = training_data)
future_dates$lm_forecast <- predict(model, newdata = future_dates)

## Logistic Regression Model
# model2 <- glm(Trip_Occurred ~ Date + 
#               price, 
#               data = training_data,
#               family = binomial)
# future_dates$logistic_forecast <- predict(model2, newdata = future_dates, type = "response")

## ARIMA Model
arima_model <- auto.arima(training_data$Trips.By.Date...Starts...Spin...Scooters,
                          xreg = training_data$price)
arima_forecast <- forecast(arima_model, 
                           h = nrow(future_dates),
                           xreg = future_dates$price)
future_dates$arima_forecast <- as.numeric(arima_forecast$mean)

## ETS Model
ets_model <- ets(training_data$Trips.By.Date...Starts...Spin...Scooters)
ets_forecast <- forecast(ets_model, h = nrow(future_dates))
future_dates$ets_forecast <- as.numeric(ets_forecast$mean)

## Prophet Model
prophet_data <- training_data %>% select(Date,
                                         Trips.By.Date...Starts...Spin...Scooters) %>% 
  rename(ds = Date, y = Trips.By.Date...Starts...Spin...Scooters)
prophet_model <- prophet(prophet_data)
future_prophet <- make_future_dataframe(prophet_model, 
                                        periods = nrow(future_dates))
prophet_forecast <- predict(prophet_model, future_prophet)
future_dates$prophet_forecast <- prophet_forecast$yhat[1:nrow(future_dates)]
```
```{r, fig.width=14, fig.height=8, message=F, warning=F}
combined_data <- actual_rides_data %>%
  full_join(future_dates, by = "Date")

price_change_1 <- as.Date("2021-07-26")
price_change_2 <- as.Date("2022-12-01")
price_change_3 <- as.Date("2024-07-01")

ggplot() +
  geom_line(data = combined_data, aes(x = Date, y = actual_rides, 
                                      color = "Actual Rides (with price change)"), 
            size = 1, alpha = 0.5) +
  geom_line(data = combined_data, aes(x = Date, y = lm_forecast, 
                                      color = "LM Forecast (no price change)"), 
            size = 1) +
  # geom_line(data = combined_data, aes(x = Date, y = logistic_forecast, 
  #                                     color = "Logistic Forecast (no price change)"), size = 1) +
  geom_line(data = combined_data, aes(x = Date, y = arima_forecast, 
                                      color = "ARIMA Forecast (no price change)"), 
            size = 1) +
  geom_line(data = combined_data, aes(x = Date, y = ets_forecast, 
                                      color = "ETS Forecast (no price change)"), 
            size = 1) +
  geom_line(data = combined_data, aes(x = Date, y = prophet_forecast, 
                                      color = "Prophet Forecast (no price change)"), 
            size = 1) +
  geom_rect(aes(xmin = as.Date("2024-07-01"),
                xmax = as.Date("2024-10-17"), 
                ymin = -Inf, ymax = Inf), 
            fill = "lightblue", alpha = 0.1) +
  labs(title = "Actual Scooter Rides vs Forecasted Scooter Rides Without Price Change (LM, ARIMA, ETS, Prophet)",
       x = "Date", 
       y = "Number of Rides",
       subtitle = "Data Collected from City of Fort Collins Spin E-Scooters") +
  scale_color_manual(values = c("Actual Rides (with price change)" = "black", 
                                "LM Forecast (no price change)" = "blue",
                                # "Logistic Forecast (no price change)" = "orange",
                                "ARIMA Forecast (no price change)" = "red", 
                                "ETS Forecast (no price change)" = "green", 
                                "Prophet Forecast (no price change)" = "purple")) +
  theme_minimal() +
  geom_vline(xintercept = price_change_1, color = "black") +
  geom_vline(xintercept = price_change_2, color = "black") +
  geom_vline(xintercept = price_change_3, color = "black") +
  annotate("text", x = as.Date("2021-07-26"), 
           y = max(combined_data$actual_rides, na.rm = TRUE),
           label = "Initial Price\n(0.30 $/hr)\nJuly 26, 2021", 
           angle = 0, hjust = -0.1) +
  annotate("text", x = as.Date("2022-12-01"), 
           y = max(combined_data$actual_rides, na.rm = TRUE),
           label = "First Price\nChange (0.40 $/hr)\nDec 1, 2022", 
           angle = 0, hjust = -0.1) +
  annotate("text", x = as.Date("2024-07-01"), 
           y = max(combined_data$actual_rides, na.rm = TRUE),
           label = "Second Price\nChange\nJuly 1, 2024", 
           angle = 0, hjust = -0.1) +
  theme(legend.title = element_blank()
)
```
