---
title: "forecasting_weekly_rides"
author: "Grace Combe"
date: "2024-11-03"
output: html_document
---

**GOAL: forecast scooter usage by week over the next year**
```{r}
set.seed(227)
```

```{r}
# load packages
library(tidyverse)
library(readr)
library(ggplot2)
library(forecast)
library(forecastHybrid)
library(fpp2)
library(TTR)
library(dplyr)
library(lubridate)
library(plotly)
library(rio)
library(tseries)
library(zoo)
library(keras)
library(tensorflow)
```

```{r}
daily_scooters <- read.csv("FOCO Spin Data/scooters 2021-07-26--2024-10-17-analyze-trips_by_date.csv")
daily_scooters <- daily_scooters %>% rename("TripsByDate" = Trips.By.Date...Starts...Spin...Scooters)
```

```{r}
daily_scooters$Date <- ymd(daily_scooters$Date)
#daily_scooters$month <- month(daily_scooters$Date)
daily_scooters$year <- year(daily_scooters$Date)
daily_scooters$WeekNum <- isoweek(ymd(daily_scooters$Date))
```

```{r}
week_data <- daily_scooters %>% group_by(WeekNum, year) %>% summarize(WeeklyRides = sum(TripsByDate)) %>% arrange(year, WeekNum)
#month_data <- daily_scooters %>% group_by(month, year) %>% summarize(MonthlyRides = sum(TripsByDate)) %>% arrange(year, month)
```


**Exploratory**
```{r}
#figure out how to connect dots?
ggplot(data = week_data, aes(x = WeekNum, y = WeeklyRides, color = as.factor(year))) +geom_point()
```




**WEEKLY DATA**
```{r}
# splitting into train/test; test is data from 2024
week_data$Class <- if_else(week_data$year == 2024, 'Test', 'Train')
train <- subset(week_data, Class == 'Train') %>% arrange(year, WeekNum)
test <- subset(week_data, Class == 'Test') %>% arrange(year, WeekNum)
nrow(train)
nrow(test)
```

```{r}
# time series
ts <- ts(train$WeeklyRides, start = c(2021, 30), frequency = 52)
summary(ts)
plot(ts)
```

```{r}
# mean absolute percentage error to measure accuracy
mape <- function(actual, pred){
mape <- mean(abs((actual-pred)/actual))*100
return(mape)
}
```

```{r}
ddata <- decompose(ts, "multiplicative")
plot(ddata)
plot(ddata$trend)
plot(ddata$seasonal)
plot(ddata$random)
plot(ts)
abline(reg = lm(ts ~time(ts)))
boxplot(ts ~ cycle(ts, xlab = "WeekNum", ylab = "WeeklyRides"))
mymodel <- auto.arima(ts)
mymodel
plot.ts(mymodel$residuals)
myforecast <- forecast(mymodel, level = c(95), h = 52)
plot(myforecast)
forecast <-data.frame(ExpApps = as.numeric(myforecast$mean),
                      Lower = as.numeric(myforecast$lower),
                      Upper = as.numeric(myforecast$upper), stringsAsFactors = F)
```

**ARIMA Model**
```{r}
arima_model <- auto.arima(ts, D = 1)
summary(arima_model)
```

```{r}
# mape for test data
fore_arima = forecast::forecast(arima_model, h = 42)
df_arima <- as.data.frame(fore_arima)
test$arima = df_arima$`Point Forecast`
mape(test$WeeklyRides, test$arima[1:42])
```
MAPE 64.95 on test data


**Graphing ARIMA Model**
```{r}
df_arima <- df_arima %>% mutate(Week = row_number(),
                    ForCalc = Week -52,
                    Year = ifelse(ForCalc > 0, 2025, 2024),
                    Week = ifelse(ForCalc > 0, ForCalc, Week),
                    Category = "Forecast") %>%
  select(Year, Week, `Point Forecast`, `Lo 95`, `Hi 95`)
```

```{r}
plot <- ggplot(test, aes(WeekNum)) +
  geom_line(aes(y = arima, color = "Forecast")) +
  geom_line(aes(y = WeeklyRides, color = "Actual"))+
  scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) +
  ggtitle("Weekly Rides 2024")
ggplotly(plot)
```

```{r}
model <- auto.arima(ts, D = 1)
forecasted_values <- forecast(model, h = 52)
plot(forecasted_values) # going negative, not good
```

**STLF**
```{r}
stlf_forecast <- stlf(ts, h = 52)
plot(stlf_forecast)
```

```{r}
df_stlf <- as.data.frame(stlf_forecast)
test$stlf = df_stlf$`Point Forecast`[1:42]
mape(test$WeeklyRides, test$stlf)
```
47.12 mape on test 

**Graphing STLF**
```{r}
df_stlf <- df_stlf %>% mutate(Week = row_number(),
                    ForCalc = Week -52,
                    Year = ifelse(ForCalc > 0, 2025, 2024),
                    Week = ifelse(ForCalc > 0, ForCalc, Week),
                    Category = "Forecast") %>%
  select(Year, Week, `Point Forecast`, `Lo 95`, `Hi 95`)
```

```{r}
plot <- ggplot(test, aes(WeekNum)) +
  geom_line(aes(y = stlf, color = "Forecast")) +
  geom_line(aes(y = WeeklyRides, color = "Actual"))+
  scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) +
  ggtitle("STLF Weekly Rides 2024")
ggplotly(plot)
```

**Another ARIMA**
```{r}
arima_forecast <- stlf(ts, h = 52, s.window = "periodic", method = "arima")
plot(arima_forecast)
```

```{r}
df_ar <- as.data.frame(arima_forecast)
test$ar = df_ar$`Point Forecast`[1:42]
mape(test$WeeklyRides, test$ar)
```
40.11 mape on test

```{r}
df_ar <- df_ar %>% mutate(Week = row_number(),
                    ForCalc = Week -52,
                    Year = ifelse(ForCalc > 0, 2025, 2024),
                    Week = ifelse(ForCalc > 0, ForCalc, Week),
                    Category = "Forecast") %>%
  select(Year, Week, `Point Forecast`, `Lo 95`, `Hi 95`)
```

```{r}
plot <- ggplot(test, aes(WeekNum)) +
  geom_line(aes(y = ar, color = "Forecast")) +
  geom_line(aes(y = WeeklyRides, color = "Actual"))+
  scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) +
  ggtitle("ARIMA Weekly Rides 2024")
ggplotly(plot)
```

**Hybrid Models**
```{r}
library(forecastHybrid)
mod1 <- hybridModel(ts)
mod1_forecast <- forecast(mod1, h = 52)
plot(mod1_forecast)
```

```{r}
df_hybrid <- as.data.frame(mod1_forecast)
test$hybrid = df_hybrid$`Point Forecast`[1:42]
mape(test$WeeklyRides, test$hybrid)
```
32.97 mape on test

```{r}
df_hybrid <- df_hybrid %>% mutate(Week = row_number(),
                    ForCalc = Week -52,
                    Year = ifelse(ForCalc > 0, 2025, 2024),
                    Week = ifelse(ForCalc > 0, ForCalc, Week),
                    Category = "Forecast") %>%
  select(Year, Week, `Point Forecast`, `Lo 95`, `Hi 95`)
```

```{r}
plot <- ggplot(test, aes(WeekNum)) +
  geom_line(aes(y = hybrid, color = "Forecast")) +
  geom_line(aes(y = WeeklyRides, color = "Actual"))+
  scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) +
  ggtitle("Hybrid Mod1 Weekly Rides 2024")
ggplotly(plot)
```

**ARIMA/STLM Hybrid**
```{r}
# ARIMA, STLM
mod_as <- hybridModel(ts, models = "as")
as_fore <- forecast(mod_as, h = 52)
plot(as_fore)
```

```{r}
df_as <- as.data.frame(as_fore)
test$as = df_as$`Point Forecast`[1:42]
mape(test$WeeklyRides, test$as)
```
54.25 mape on test

```{r}
df_as <- df_as %>% mutate(Week = row_number(),
                    ForCalc = Week -52,
                    Year = ifelse(ForCalc > 0, 2025, 2024),
                    Week = ifelse(ForCalc > 0, ForCalc, Week),
                    Category = "Forecast") %>% 
  select(Year, Week, `Point Forecast`, `Lo 95`, `Hi 95`)
```

```{r}
#test$AppsForecast <- left_join(test, df_as, by = c("Year"= "Year", "Week"="Week"))  %>% rename("as" = "Point Forecast")
```

```{r}
plot <- ggplot(test, aes(WeekNum)) +
  geom_line(aes(y = as, color = "Forecast")) +
  geom_line(aes(y = WeeklyRides, color = "Actual"))+
  scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) + 
  ggtitle("ARIMA + STLM Num Apps 2024")
ggplotly(plot)
```

**ARIMA/THETA/NNETAR
```{r}
# ARIMA, THETA, NNETAR
mod_afn <- hybridModel(ts, models = "afn")
afn_fore <- forecast(mod_afn, h = 52)
plot(afn_fore)
```

```{r}
df_afn <- as.data.frame(afn_fore)
test$afn = df_afn$`Point Forecast`[1:42]
mape(test$WeeklyRides, test$afn)
```
31.44 mape on test

```{r}
df_afn <- df_afn %>% mutate(Week = row_number(),
                    ForCalc = Week -52,
                    Year = ifelse(ForCalc > 0, 2025, 2024),
                    Week = ifelse(ForCalc > 0, ForCalc, Week),
                    Category = "Forecast") %>%
  select(Year, Week, `Point Forecast`, `Lo 95`, `Hi 95`)
```

```{r}
#joined_data <- left_join(joined_data, df_afn, by = c("Year"= "Year", "Week"="Week"))  %>% rename("afn" = "Point Forecast")
```

```{r}
plot <- ggplot(test, aes(WeekNum)) +
  geom_line(aes(y = afn, color = "Forecast")) +
  geom_line(aes(y = WeeklyRides, color = "Actual"))+
  scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) +
  ggtitle("ARIMA + THETAM + NNETAR Weekly Rides 2024")
ggplotly(plot)
```

**ARIMA/STLM/THETAM**
```{r}
# ARIMA, STLM, THETAM
mod_afs <- hybridModel(ts, models = "afs")
afs_fore <- forecast(mod_afs, h = 52)
plot(afs_fore)
```

```{r}
df_afs <- as.data.frame(afs_fore)
test$afs = df_afs$`Point Forecast`[1:42]
mape(test$WeeklyRides, test$afs)
```
29.69 mape on test

```{r}
df_afs <- df_afs %>% mutate(Week = row_number(),
                    ForCalc = Week -52,
                    Year = ifelse(ForCalc > 0, 2025, 2024),
                    Week = ifelse(ForCalc > 0, ForCalc, Week),
                    Category = "Forecast") %>%
  select(Year, Week, `Point Forecast`, `Lo 95`, `Hi 95`)
```

```{r}
#joined_data <- left_join(joined_data, df_afs, by = c("Year"= "Year", "Week"="Week"))  %>% rename("afs" = "Point Forecast")
```

```{r}
plot <- ggplot(test, aes(WeekNum)) +
  geom_line(aes(y = afs, color = "Forecast")) +
  geom_line(aes(y = WeeklyRides, color = "Actual"))+
  scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) +
  ggtitle("AFS WeeklyRides 2024")
ggplotly(plot)
```


**Auto ARIMA**
Exploratory
```{r}
train_ts <- ts(train$WeeklyRides)
train$clean_count <- tsclean(train_ts) # cleans missing data, really bad outliers, etc
```

```{r}
# plot cleaned data; not as many outliers
#ggplot() + geom_line(data = train, aes(x = WeekEndingDate, y = clean_count)) + ylab('Cleaned Count Apps')
# plot uncleaned data
#ggplot() + geom_line(data = train, aes(x = WeekEndingDate, y = numApps)) + ylab('Uncleaned Count Apps')
```

```{r}
# moving averages
train$count_ma = ma(train$clean_count, order = 7)
train$count_ma30 = ma(train$clean_count, order = 30)
train <- replace(train, TRUE, lapply(train, na.aggregate)) #fills in NAs with column mean
```

```{r}
# plot with NAs removed
# ggplot() +
#   geom_line(data = train, aes(x = WeekEndingDate, y = clean_count, colour = "Counts")) +
#   geom_line(data = train, aes(x = WeekEndingDate, y = count_ma, colour = "Weekly Moving Average")) +
#   geom_line(data = train, aes(x = WeekEndingDate, y = count_ma30, colour = "Monthly Moving Average")) +
#   ylab('Cleaned Count Apps')
```

Decomposition
```{r}
count_ma <- ts(na.omit(train$clean_count), frequency = 52)
decomp <- stl(count_ma, "periodic")
deseasonal_count <- seasadj(decomp) #used in ARIMA model later
plot(decomp)
# clear seasonality
```

```{r}
# test for stationarity
adf.test(count_ma, alternative = "stationary")
# p value of less than 0.05 indicates stationary data set; 0.07
```


```{r}
# auto correlation and seasonal differencing
# ACF plots display correlation between a series and its lags; don't want to go outside dotted lines, if so = lags
Acf(count_ma, main = '') # outside lines
# PACF plots display correlation between a series and its lags that explained by previous lags
Pacf(count_ma, main = '') # only one outside lines
```

```{r}
# use d = 1 if in bounds, d = 2 if out of bounds (try to use 1 unless you can't)
count_d2 <- diff(deseasonal_count, differences = 2)
plot(count_d2)
```

```{r}
# retest with adf
adf.test(count_d2, alternative = "stationary") # yay less than 0.05 now; we want to be as stationary as possible
```

```{r}
# look for spikes at specific lag points of the differenced series (using count_d2 from above)
Acf(count_d2, main = 'ACF for Differenced Series') # biggest lag at 1
Pacf(count_d2, main = 'PACF for Differenced Series') # biggest lag at 1 again
```

Fitting ARIMA model
```{r}
# get auto fit p, d, q models
auto.arima(deseasonal_count, seasonal = FALSE)
```

```{r}
# evaluate and iterate - does the model make sense?
fit <- auto.arima(deseasonal_count, seasonal = FALSE)
tsdisplay(residuals(fit), lag.max = 52, main = '(2,1,1) Model Residuals') # looks like biggest lag at 52
```

```{r}
# adjust for diff = 2, lag at 52
fit2 <- arima(deseasonal_count, order = c(0,2,52))
tsdisplay(residuals(fit2), lag.max = 52, main = 'Seasonal Model Residuals')
```

```{r}
train$deseasonal_count <- deseasonal_count
```

```{r}
# testing
fit2_forecast <- data.frame(forecast(fit2, h = 42))
test$fit2 <- fit2_forecast$Point.Forecast
```

```{r}
mape(test$WeeklyRides, test$fit2)
```
113.06 mape on test

```{r}
plot <- ggplot(test, aes(WeekNum)) +
  geom_line(aes(y = fit2, color = "Forecast")) +
  geom_line(aes(y = WeeklyRides, color = "Actual"))+
  scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) +
  ggtitle("Fit2 ARIMA Weekly Rides 2024")
ggplotly(plot)
```


```{r}
# bring back seasonality?
seas_fit <- auto.arima(deseasonal_count, seasonal = TRUE)
seas_forecast <- forecast(seas_fit, h = 42)
plot(seas_forecast) # straight line is bad
```

Further testing
```{r}
# adjust for diff = 2, lag at 52
fit3 <- arima(deseasonal_count, order = c(1,2,52))
tsdisplay(residuals(fit3), lag.max = 52, main = 'Seasonal Model Residuals') # lags look the same
```

```{r}
# testing
fit3_forecast <- data.frame(forecast(fit3, h = 42))
test$fit3 <- fit3_forecast$Point.Forecast
```

```{r}
mape(test$WeeklyRides, test$fit3)
```
112.29 mape on test

```{r}
plot <- ggplot(test, aes(WeekNum)) +
  geom_line(aes(y = fit3, color = "Forecast")) +
  geom_line(aes(y = WeeklyRides, color = "Actual"))+
  scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) +
  ggtitle("Fit3 ARIMA Weekly Rides 2024")
ggplotly(plot)
```


**REFITTING MODELS WITH CLEANED DATA**

```{r}
ts_clean <- ts(train$clean_count, start = c(2021, 1), frequency = 52)
```

```{r}
ddata <- decompose(ts_clean, "multiplicative")
plot(ddata)
plot(ddata$trend)
plot(ddata$seasonal)
plot(ddata$random)
plot(ts_clean)
abline(reg = lm(ts_clean ~time(ts_clean)))
boxplot(ts_clean ~ cycle(ts_clean, xlab = "WeekOfYear", ylab = "numApps"))
mymodel <- auto.arima(ts_clean)
mymodel
plot.ts(mymodel$residuals)
myforecast <- forecast(mymodel, level = c(95), h = 52)
plot(myforecast)
forecast <-data.frame(ExpApps = as.numeric(myforecast$mean),
                      Lower = as.numeric(myforecast$lower),
                      Upper = as.numeric(myforecast$upper), stringsAsFactors = F)
```

**ARIMA Model**
```{r}
arima_model <- auto.arima(ts_clean, D = 1)
summary(arima_model)
```

```{r}
# mape for test data
fore_arima = forecast::forecast(arima_model, h = 42)
df_arima <- as.data.frame(fore_arima)
test$arima = df_arima$`Point Forecast`
mape(test$WeeklyRides, test$arima[1:42])
```
66.40 mape on test

**Graphing ARIMA Model**
```{r}
df_arima <- df_arima %>% mutate(Week = row_number(),
                    ForCalc = Week -52,
                    Year = ifelse(ForCalc > 0, 2025, 2024),
                    Week = ifelse(ForCalc > 0, ForCalc, Week),
                    Category = "Forecast") %>%
  select(Year, Week, `Point Forecast`, `Lo 95`, `Hi 95`)
```

```{r}
#joined_data <- left_join(df_arima, test, by = c("Year"= "Year", "Week"="WeekNum")) %>% rename("arima1_clean" = "Point Forecast")
```

```{r}
plot <- ggplot(test, aes(WeekNum)) +
  geom_line(aes(y = arima, color = "Forecast")) +
  geom_line(aes(y = WeeklyRides, color = "Actual"))+
  scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) +
  ggtitle("Weekly Rides 2024")
ggplotly(plot)
```

**Different ARIMA**
```{r}
model <- auto.arima(ts_clean, D = 1)
forecasted_values <- forecast(model, h = 52)
plot(forecasted_values)
```

**STLF on clean**
```{r}
stlf_forecast <- stlf(ts_clean, h = 52)
plot(stlf_forecast)
```

```{r}
df_stlf <- as.data.frame(stlf_forecast)
test$stlf_clean = df_stlf$`Point Forecast`[1:42]
mape(test$WeeklyRides, test$stlf_clean)
```
46.39 mape on test

**Graphing STLF**
```{r}
df_stlf <- df_stlf %>% mutate(Week = row_number(),
                    ForCalc = Week -52,
                    Year = ifelse(ForCalc > 0, 2025, 2024),
                    Week = ifelse(ForCalc > 0, ForCalc, Week),
                    Category = "Forecast") %>%
  select(Year, Week, `Point Forecast`, `Lo 95`, `Hi 95`)
```

```{r}
#joined_data <- left_join(joined_data, df_stlf, by = c("Year"= "Year", "Week"="Week")) %>% rename("stlf_clean" = "Point Forecast")
```

```{r}
plot <- ggplot(test, aes(WeekNum)) +
  geom_line(aes(y = stlf_clean, color = "Forecast")) +
  geom_line(aes(y = WeeklyRides, color = "Actual"))+
  scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) +
  ggtitle("STLF Weekly Rides 2024")
ggplotly(plot)
```


```{r}
arima_forecast <- stlf(ts_clean, h = 52, s.window = "periodic", method = "arima")
plot(arima_forecast)
```

```{r}
df_ar <- as.data.frame(arima_forecast)
test$ar_clean = df_ar$`Point Forecast`[1:42]
mape(test$WeeklyRides, test$ar_clean)
```
42.31 mape on test

```{r}
df_ar <- df_ar %>% mutate(Week = row_number(),
                    ForCalc = Week -52,
                    Year = ifelse(ForCalc > 0, 2025, 2024),
                    Week = ifelse(ForCalc > 0, ForCalc, Week),
                    Category = "Forecast") %>%
  select(Year, Week, `Point Forecast`, `Lo 95`, `Hi 95`)
```

```{r}
#joined_data <- left_join(joined_data, df_ar, by = c("Year"= "Year", "Week"="Week"))  %>% rename("arima2" = "Point Forecast")
```

```{r}
plot <- ggplot(test, aes(WeekNum)) +
  geom_line(aes(y = ar_clean, color = "Forecast")) +
  geom_line(aes(y = WeeklyRides, color = "Actual"))+
  scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) +
  ggtitle("ARIMA Weekly Rides 2024")
ggplotly(plot)
```

**Hybrid Models**
```{r}
library(forecastHybrid)
mod1 <- hybridModel(ts_clean)
mod1_forecast <- forecast(mod1, h = 52)
plot(mod1_forecast)
```

```{r}
df_hybrid2 <- as.data.frame(mod1_forecast)
test$hybrid2 = df_hybrid2$`Point Forecast`[1:42]
mape(test$WeeklyRides, test$hybrid2)
```
33.58 mape on test

```{r}
df_hybrid2 <- df_hybrid2 %>% mutate(Week = row_number(),
                    ForCalc = Week -52,
                    Year = ifelse(ForCalc > 0, 2025, 2024),
                    Week = ifelse(ForCalc > 0, ForCalc, Week),
                    Category = "Forecast") %>%
  select(Year, Week, `Point Forecast`, `Lo 95`, `Hi 95`)
```

```{r}
#joined_data <- left_join(joined_data, df_hybrid2, by = c("Year"= "Year", "Week"="Week"))  %>% rename("hybrid2" = "Point Forecast")
```

```{r}
plot <- ggplot(test, aes(WeekNum)) +
  geom_line(aes(y = hybrid2, color = "Forecast")) +
  geom_line(aes(y = WeeklyRides, color = "Actual"))+
  scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) +
  ggtitle("Hybrid2 Weekly Rides 2024")
ggplotly(plot)
```


**ARIMA/STLM**
```{r}
# ARIMA, STLM
mod_as_clean <- hybridModel(ts_clean, models = "as") # no ets bc h >24
as_fore_clean <- forecast(mod_as_clean, h = 52)
plot(as_fore_clean)
```

```{r}
df_as_clean <- as.data.frame(as_fore_clean)
test$as_clean = df_as$`Point Forecast`[1:42]
mape(test$WeeklyRides, test$as_clean)
```
54.25 mape on test

```{r}
df_as_clean <- df_as_clean %>% mutate(Week = row_number(),
                    ForCalc = Week -52,
                    Year = ifelse(ForCalc > 0, 2025, 2024),
                    Week = ifelse(ForCalc > 0, ForCalc, Week),
                    Category = "Forecast") %>%
  select(Year, Week, `Point Forecast`, `Lo 95`, `Hi 95`)
```

```{r}
#joined_data <- left_join(joined_data, df_as_clean, by = c("Year"= "Year", "Week"="Week"))  %>% rename("as2_clean" = "Point Forecast")
```

```{r}
plot <- ggplot(test, aes(WeekNum)) +
  geom_line(aes(y = as_clean, color = "Forecast")) +
  geom_line(aes(y = WeeklyRides, color = "Actual"))+
  scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) +
  ggtitle("ARIMA + STLM Weekly Rides 2024")
ggplotly(plot)
```









