---
title: "forecasting_monthly_rides"
author: "Grace Combe"
date: "2024-11-03"
output: html_document
---

**GOAL: forecast scooter usage by month over the next year**
```{r}
set.seed(227)
```


```{r}
# load packages
library(tidyverse)
library(readr)
library(ggplot2)
library(forecast)
library(forecastHybrid)
library(fpp2)
library(TTR)
library(dplyr)
library(lubridate)
library(plotly)
library(rio)
library(tseries)
library(zoo)
library(keras)
library(tensorflow)
```

```{r}
daily_scooters <- read.csv("FOCO Spin Data/scooters 2021-07-26--2024-10-17-analyze-trips_by_date.csv")
daily_scooters <- daily_scooters %>% rename("TripsByDate" = Trips.By.Date...Starts...Spin...Scooters)
```

```{r}
daily_scooters$Date <- ymd(daily_scooters$Date)
daily_scooters$month <- month(daily_scooters$Date)
daily_scooters$year <- year(daily_scooters$Date)
#daily_scooters$WeekNum <- isoweek(ymd(daily_scooters$Date))
```

```{r}
#week_data <- daily_scooters %>% group_by(WeekNum, year) %>% summarize(WeeklyRides = sum(TripsByDate)) %>% arrange(year, WeekNum)
month_data <- daily_scooters %>% group_by(month, year) %>% summarize(MonthlyRides = sum(TripsByDate)) %>% arrange(year, month)
```

**Exploratory**
```{r}
#figure out how to connect dots?
ggplot(data = month_data, aes(x = month, y = MonthlyRides, color = as.factor(year))) +geom_point()
```

```{r}
# month_data <- month_data %>% mutate(adjPrice = case_when(
#   row_number() <= 17 ~ MonthlyRides/0.30,
#   row_number() > 17 & row_number() <= 36 ~ MonthlyRides/0.40,
#   row_number() > 36 ~ MonthlyRides/0.30
#   )
# )
```

```{r}
#month_data <- month_data %>% rename("OrigRides" = "MonthlyRides", "MonthlyRides" = "adjPrice")
```


**MONTHLY DATA**
```{r}
# splitting into train/test; test is data from 2024
month_data$Class <- if_else(month_data$year == 2024, 'Test', 'Train')
train <- subset(month_data, Class == 'Train') %>% arrange(year, month)
test <- subset(month_data, Class == 'Test') %>% arrange(year, month)
nrow(train)
nrow(test)
```

```{r}
#train <- train %>% filter(year >= 2022) %>% arrange(year, month)
```


```{r}
# time series
ts <- ts(train$MonthlyRides, start = c(2021, 7), frequency = 12)
summary(ts)
plot(ts)
```

```{r}
# Create time series object
ts_data <- ts(train$MonthlyRides, start = c(2021, 7), frequency = 12)

# Custom weights: Assign higher weight to the first 6 months
weights <- c(rep(2, 6), rep(1, length(ts_data) - 6))  # Weight 2 for the first 6 months, weight 1 for the rest

# Compute weighted average (use this as a simple example)
weighted_avg <- sum(ts_data * weights) / sum(weights)

# Print the weighted average
print(weighted_avg)
```


```{r}
# mean absolute percentage error to measure accuracy
mape <- function(actual, pred){
mape <- mean(abs((actual-pred)/actual))*100
return(mape)
}
```

```{r}
ddata <- decompose(ts, "multiplicative")
plot(ddata)
plot(ddata$trend)
plot(ddata$seasonal)
plot(ddata$random)
plot(ts)
abline(reg = lm(ts ~time(ts)))
boxplot(ts ~ cycle(ts, xlab = "month", ylab = "MonthlyRides"))
mymodel <- auto.arima(ts)
mymodel
plot.ts(mymodel$residuals)
myforecast <- forecast(mymodel, level = c(95), h = 12)
plot(myforecast)
forecast <-data.frame(ExpApps = as.numeric(myforecast$mean),
                      Lower = as.numeric(myforecast$lower),
                      Upper = as.numeric(myforecast$upper), stringsAsFactors = F)
```

**ARIMA Model**
```{r}
arima_model <- auto.arima(ts, D = 1)
summary(arima_model)
```

```{r}
# mape for test data
fore_arima = forecast::forecast(arima_model, h = 10)
df_arima <- as.data.frame(fore_arima)
test$arima = df_arima$`Point Forecast`
mape(test$MonthlyRides, test$arima[1:10])
```
MAPE 34.33 on test data (with 2021)
37.87 (without 2021)


**Graphing ARIMA Model**
```{r}
df_arima <- df_arima %>% mutate(Month = row_number(),
                    ForCalc = Month -12,
                    Year = ifelse(ForCalc > 0, 2025, 2024),
                    Month = ifelse(ForCalc > 0, ForCalc, Month),
                    Category = "Forecast") %>%
  select(Year, Month, `Point Forecast`, `Lo 95`, `Hi 95`)
```

```{r}
plot <- ggplot(test, aes(month)) +
  geom_line(aes(y = arima, color = "Forecast")) +
  geom_line(aes(y = MonthlyRides, color = "Actual"))+
  scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) +
  ggtitle("Monthly Rides 2024")
ggplotly(plot)
```

```{r}
model <- auto.arima(ts, D = 1)
forecasted_values <- forecast(model, h = 12)
plot(forecasted_values)
```

**STLF**
```{r}
stlf_forecast <- stlf(ts, h = 12)
plot(stlf_forecast)
```

```{r}
df_stlf <- as.data.frame(stlf_forecast)
test$stlf = df_stlf$`Point Forecast`[1:10]
mape(test$MonthlyRides, test$stlf)
```
36.49 mape on test 

**Graphing STLF**
```{r}
df_stlf <- df_stlf %>% mutate(Month = row_number(),
                    ForCalc = Month -12,
                    Year = ifelse(ForCalc > 0, 2025, 2024),
                    Month = ifelse(ForCalc > 0, ForCalc, Month),
                    Category = "Forecast") %>%
  select(Year, Month, `Point Forecast`, `Lo 95`, `Hi 95`)
```

```{r}
plot <- ggplot(test, aes(month)) +
  geom_line(aes(y = stlf, color = "Forecast")) +
  geom_line(aes(y = MonthlyRides, color = "Actual"))+
  scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) +
  ggtitle("STLF Monthly Rides 2024")
ggplotly(plot)
```

**Another ARIMA**
```{r}
arima_forecast <- stlf(ts, h = 12, s.window = "periodic", method = "arima")
plot(arima_forecast)
```

```{r}
df_ar <- as.data.frame(arima_forecast)
test$ar = df_ar$`Point Forecast`[1:10]
mape(test$MonthlyRides, test$ar)
```
29.77 mape on test

```{r}
df_ar <- df_ar %>% mutate(Month = row_number(),
                    ForCalc = Month -12,
                    Year = ifelse(ForCalc > 0, 2025, 2024),
                    Month = ifelse(ForCalc > 0, ForCalc, Month),
                    Category = "Forecast") %>%
  select(Year, Month, `Point Forecast`, `Lo 95`, `Hi 95`)
```

```{r}
plot <- ggplot(test, aes(month)) +
  geom_line(aes(y = ar, color = "Forecast")) +
  geom_line(aes(y = MonthlyRides, color = "Actual"))+
  scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) +
  ggtitle("ARIMA Monthly Rides 2024")
ggplotly(plot)
```

**Hybrid Models**
```{r}
library(forecastHybrid)
mod1 <- hybridModel(ts)
mod1_forecast <- forecast(mod1, h = 12)
plot(mod1_forecast)
```

```{r}
df_hybrid <- as.data.frame(mod1_forecast)
test$hybrid = df_hybrid$`Point Forecast`[1:10]
mape(test$MonthlyRides, test$hybrid)
```
9.73 mape on test

```{r}
df_hybrid <- df_hybrid %>% mutate(Month = row_number(),
                    ForCalc = Month -12,
                    Year = ifelse(ForCalc > 0, 2025, 2024),
                    Month = ifelse(ForCalc > 0, ForCalc, Month),
                    Category = "Forecast") %>%
  select(Year, Month, `Point Forecast`, `Lo 95`, `Hi 95`)
```

```{r}
plot <- ggplot(test, aes(month)) +
  geom_line(aes(y = hybrid, color = "Forecast")) +
  geom_line(aes(y = MonthlyRides, color = "Actual"))+
  scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) +
  ggtitle("Hybrid Mod1 Monthly Rides 2024")
ggplotly(plot)
```

**ARIMA/STLM Hybrid**
```{r}
# ARIMA, STLM
mod_as <- hybridModel(ts, models = "as")
as_fore <- forecast(mod_as, h = 12)
plot(as_fore)
```

```{r}
df_as <- as.data.frame(as_fore)
test$as = df_as$`Point Forecast`[1:10]
mape(test$MonthlyRides, test$as)
```
21.08 mape on test

```{r}
df_as <- df_as %>% mutate(Month = row_number(),
                    ForCalc = Month -12,
                    Year = ifelse(ForCalc > 0, 2025, 2024),
                    Month = ifelse(ForCalc > 0, ForCalc, Month),
                    Category = "Forecast") %>% 
  select(Year, Month, `Point Forecast`, `Lo 95`, `Hi 95`)
```

```{r}
#test$AppsForecast <- left_join(test, df_as, by = c("Year"= "Year", "Week"="Week"))  %>% rename("as" = "Point Forecast")
```

```{r}
plot <- ggplot(test, aes(month)) +
  geom_line(aes(y = as, color = "Forecast")) +
  geom_line(aes(y = MonthlyRides, color = "Actual"))+
  scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) + 
  ggtitle("ARIMA + STLM Num Apps 2024")
ggplotly(plot)
```

**ARIMA/STLM/ETS Hybrid**
```{r}
# ARIMA, STLM
mod_aes <- hybridModel(ts, models = "aes")
aes_fore <- forecast(mod_aes, h = 12)
plot(aes_fore)
```

```{r}
df_aes <- as.data.frame(aes_fore)
test$aes = df_aes$`Point Forecast`[1:10]
mape(test$MonthlyRides, test$aes)
```
17.457 mape on test

```{r}
df_as <- df_as %>% mutate(Month = row_number(),
                    ForCalc = Month -12,
                    Year = ifelse(ForCalc > 0, 2025, 2024),
                    Month = ifelse(ForCalc > 0, ForCalc, Month),
                    Category = "Forecast") %>% 
  select(Year, Month, `Point Forecast`, `Lo 95`, `Hi 95`)
```

```{r}
#test$AppsForecast <- left_join(test, df_as, by = c("Year"= "Year", "Week"="Week"))  %>% rename("as" = "Point Forecast")
```

```{r}
plot <- ggplot(test, aes(month)) +
  geom_line(aes(y = aes, color = "Forecast")) +
  geom_line(aes(y = MonthlyRides, color = "Actual"))+
  scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) + 
  ggtitle("ARIMA + STLM + ETS Num Apps 2024")
ggplotly(plot)
```

**ARIMA/STLM/ETS/TBATS Hybrid**
```{r}
# ARIMA, STLM
mod_aets <- hybridModel(ts, models = "aets")
aets_fore <- forecast(mod_aets, h = 12)
plot(aets_fore)
```

```{r}
df_aets <- as.data.frame(aets_fore)
test$aets = df_aets$`Point Forecast`[1:10]
mape(test$MonthlyRides, test$aets)
```
17.631 mape on test

```{r}
df_aets <- df_aets %>% mutate(Month = row_number(),
                    ForCalc = Month -12,
                    Year = ifelse(ForCalc > 0, 2025, 2024),
                    Month = ifelse(ForCalc > 0, ForCalc, Month),
                    Category = "Forecast") %>% 
  select(Year, Month, `Point Forecast`, `Lo 95`, `Hi 95`)
```

```{r}
#test$AppsForecast <- left_join(test, df_as, by = c("Year"= "Year", "Week"="Week"))  %>% rename("as" = "Point Forecast")
```

```{r}
plot <- ggplot(test, aes(month)) +
  geom_line(aes(y = aets, color = "Forecast")) +
  geom_line(aes(y = MonthlyRides, color = "Actual"))+
  scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) + 
  ggtitle("ARIMA + STLM +ETS +TBATS Num Apps 2024")
ggplotly(plot)
```




**ARIMA/THETA/NNETAR**
```{r}
# ARIMA, THETA, NNETAR
mod_afn <- hybridModel(ts, models = "afn")
afn_fore <- forecast(mod_afn, h = 12)
plot(afn_fore)
```

```{r}
df_afn <- as.data.frame(afn_fore)
test$afn = df_afn$`Point Forecast`[1:10]
mape(test$MonthlyRides, test$afn)
```
26.09 mape on test

```{r}
df_afn <- df_afn %>% mutate(Month = row_number(),
                    ForCalc = Month -12,
                    Year = ifelse(ForCalc > 0, 2025, 2024),
                    Month = ifelse(ForCalc > 0, ForCalc, Month),
                    Category = "Forecast") %>%
  select(Year, Month, `Point Forecast`, `Lo 95`, `Hi 95`)
```

```{r}
#joined_data <- left_join(joined_data, df_afn, by = c("Year"= "Year", "Week"="Week"))  %>% rename("afn" = "Point Forecast")
```

```{r}
plot <- ggplot(test, aes(month)) +
  geom_line(aes(y = afn, color = "Forecast")) +
  geom_line(aes(y = MonthlyRides, color = "Actual"))+
  scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) +
  ggtitle("ARIMA + THETAM + NNETAR Monthly Rides 2024")
ggplotly(plot)
```

**ARIMA/STLM/THETAM**
```{r}
# ARIMA, STLM, THETAM
mod_afs <- hybridModel(ts, models = "afs")
afs_fore <- forecast(mod_afs, h = 12)
plot(afs_fore)
```

```{r}
df_afs <- as.data.frame(afs_fore)
test$afs = df_afs$`Point Forecast`[1:10]
mape(test$MonthlyRides, test$afs)
```
19.18 on test

```{r}
df_afs <- df_afs %>% mutate(Month = row_number(),
                    ForCalc = Month -12,
                    Year = ifelse(ForCalc > 0, 2025, 2024),
                    Month = ifelse(ForCalc > 0, ForCalc, Month),
                    Category = "Forecast") %>%
  select(Year, Month, `Point Forecast`, `Lo 95`, `Hi 95`)
```

```{r}
#joined_data <- left_join(joined_data, df_afs, by = c("Year"= "Year", "Week"="Week"))  %>% rename("afs" = "Point Forecast")
```

```{r}
plot <- ggplot(test, aes(month)) +
  geom_line(aes(y = afs, color = "Forecast")) +
  geom_line(aes(y = MonthlyRides, color = "Actual"))+
  scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) +
  ggtitle("AFS MonthlyRides 2024")
ggplotly(plot)
```


**Auto ARIMA**
Exploratory
```{r}
train_ts <- ts(train$MonthlyRides)
train$clean_count <- tsclean(train_ts) # cleans missing data, really bad outliers, etc
```

```{r}
# plot cleaned data; not as many outliers
ggplot() + geom_line(data = train, aes(x = month, y = clean_count)) + ylab('Cleaned Count Rides')
# plot uncleaned data
ggplot() + geom_line(data = train, aes(x = month, y = MonthlyRides)) + ylab('Uncleaned Count Rides')
```

```{r}
# moving averages
train$count_ma = ma(train$clean_count, order = 7)
#train$count_ma30 = ma(train$clean_count, order = )
#train <- replace(train, TRUE, lapply(train, na.aggregate)) #fills in NAs with column mean
```

```{r}
# plot with NAs removed
# ggplot() +
#   geom_line(data = train, aes(x = WeekEndingDate, y = clean_count, colour = "Counts")) +
#   geom_line(data = train, aes(x = WeekEndingDate, y = count_ma, colour = "Weekly Moving Average")) +
#   geom_line(data = train, aes(x = WeekEndingDate, y = count_ma30, colour = "Monthly Moving Average")) +
#   ylab('Cleaned Count Apps')
```

Decomposition
```{r}
count_ma <- ts(na.omit(train$clean_count), frequency = 12)
decomp <- stl(count_ma, "periodic")
deseasonal_count <- seasadj(decomp) #used in ARIMA model later
plot(decomp)
# clear seasonality
```

```{r}
# test for stationarity
adf.test(count_ma, alternative = "stationary")
# p value of less than 0.05 indicates stationary data set
```
0.3861 -> not stationary

```{r}
# auto correlation and seasonal differencing
# ACF plots display correlation between a series and its lags; don't want to go outside dotted lines, if so = lags
Acf(count_ma, main = '') # outside lines
# PACF plots display correlation between a series and its lags that explained by previous lags
Pacf(count_ma, main = '') # only two outside lines
```

```{r}
# use d = 1 if in bounds, d = 2 if out of bounds (try to use 1 unless you can't)
count_d2 <- diff(deseasonal_count, differences = 2)
plot(count_d2)
```

```{r}
# retest with adf
adf.test(count_d2, alternative = "stationary") # yay less than 0.05 now; we want to be as stationary as possible
```

```{r}
# look for spikes at specific lag points of the differenced series (using count_d2 from above)
Acf(count_d2, main = 'ACF for Differenced Series') # all within bounds
Pacf(count_d2, main = 'PACF for Differenced Series') # all within bounds
```

Fitting ARIMA model
```{r}
# get auto fit p, d, q models
auto.arima(deseasonal_count, seasonal = FALSE)
```

```{r}
# evaluate and iterate - does the model make sense?
fit <- auto.arima(deseasonal_count, seasonal = FALSE)
tsdisplay(residuals(fit), lag.max = 30, main = '(0,0,1) Model Residuals') # looks like biggest lag at 24, still in bounds
```

```{r}
# adjust for diff = 2, lag at 12
fit2 <- arima(deseasonal_count, order = c(0,2,12))
tsdisplay(residuals(fit2), lag.max = 30, main = 'Seasonal Model Residuals')
```

```{r}
train$deseasonal_count <- deseasonal_count
```

```{r}
# testing
fit2_forecast <- data.frame(forecast(fit2, h = 10))
test$fit2 <- fit2_forecast$Point.Forecast
```

```{r}
mape(test$MonthlyRides, test$fit2)
```
83.01 mape on test

```{r}
plot <- ggplot(test, aes(month)) +
  geom_line(aes(y = fit2, color = "Forecast")) +
  geom_line(aes(y = MonthlyRides, color = "Actual"))+
  scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) +
  ggtitle("Fit2 ARIMA Monthly Rides 2024")
ggplotly(plot)
```


```{r}
# bring back seasonality?
seas_fit <- auto.arima(deseasonal_count, seasonal = TRUE)
seas_forecast <- forecast(seas_fit, h = 10)
plot(seas_forecast) # straight line is bad
```

Further testing
```{r}
# adjust for diff = 2, lag at 12
fit3 <- arima(deseasonal_count, order = c(1,2,12))
tsdisplay(residuals(fit3), lag.max = 104, main = 'Seasonal Model Residuals') # lags look the same
```

```{r}
# testing
fit3_forecast <- data.frame(forecast(fit3, h = 10))
test$fit3 <- fit3_forecast$Point.Forecast
```

```{r}
mape(test$MonthlyRides, test$fit3)
```

```{r}
plot <- ggplot(test, aes(month)) +
  geom_line(aes(y = fit3, color = "Forecast")) +
  geom_line(aes(y = MonthlyRides, color = "Actual"))+
  scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) +
  ggtitle("Fit3 ARIMA Monthly Rides 2024")
ggplotly(plot)
```


**REFITTING MODELS WITH CLEANED DATA**

```{r}
ts_clean <- ts(train$clean_count, start = c(2021, 1), frequency = 12)
```

```{r}
ddata <- decompose(ts_clean, "multiplicative")
plot(ddata)
plot(ddata$trend)
plot(ddata$seasonal)
plot(ddata$random)
plot(ts_clean)
abline(reg = lm(ts_clean ~time(ts_clean)))
boxplot(ts_clean ~ cycle(ts_clean, xlab = "MonthOfYear", ylab = "numApps"))
mymodel <- auto.arima(ts_clean)
mymodel
plot.ts(mymodel$residuals)
myforecast <- forecast(mymodel, level = c(95), h = 12)
plot(myforecast)
forecast <-data.frame(ExpApps = as.numeric(myforecast$mean),
                      Lower = as.numeric(myforecast$lower),
                      Upper = as.numeric(myforecast$upper), stringsAsFactors = F)
```

**ARIMA Model**
```{r}
arima_model <- auto.arima(ts_clean, D = 1)
summary(arima_model)
```

```{r}
# mape for test data
fore_arima = forecast::forecast(arima_model, h = 10)
df_arima <- as.data.frame(fore_arima)
test$arima = df_arima$`Point Forecast`
mape(test$numApps, test$arima[1:10])
```

**Graphing ARIMA Model**
```{r}
df_arima <- df_arima %>% mutate(Month = row_number(),
                    ForCalc = Month -12,
                    Year = ifelse(ForCalc > 0, 2025, 2024),
                    Month = ifelse(ForCalc > 0, ForCalc, Month),
                    Category = "Forecast") %>%
  select(Year, Month, `Point Forecast`, `Lo 95`, `Hi 95`)
```

```{r}
#joined_data <- left_join(df_arima, test, by = c("Year"= "Year", "Week"="WeekNum")) %>% rename("arima1_clean" = "Point Forecast")
```

```{r}
plot <- ggplot(test, aes(month)) +
  geom_line(aes(y = arima1_clean, color = "Forecast")) +
  geom_line(aes(y = MonthlyRides, color = "Actual"))+
  scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) +
  ggtitle("Monthly Rides 2024")
ggplotly(plot)
```

**Different ARIMA**
```{r}
model <- auto.arima(ts_clean, D = 1)
forecasted_values <- forecast(model, h = 12)
plot(forecasted_values)
```

**STLF on clean**
```{r}
stlf_forecast <- stlf(ts_clean, h = 12)
plot(stlf_forecast)
```

```{r}
df_stlf <- as.data.frame(stlf_forecast)
test$stlf_clean = df_stlf$`Point Forecast`[1:10]
mape(test$MonthlyRides, test$stlf_clean)
```

**Graphing STLF**
```{r}
df_stlf <- df_stlf %>% mutate(Month = row_number(),
                    ForCalc = Month -12,
                    Year = ifelse(ForCalc > 0, 2025, 2024),
                    Month = ifelse(ForCalc > 0, ForCalc, Month),
                    Category = "Forecast") %>%
  select(Year, Month, `Point Forecast`, `Lo 95`, `Hi 95`)
```

```{r}
#joined_data <- left_join(joined_data, df_stlf, by = c("Year"= "Year", "Week"="Week")) %>% rename("stlf_clean" = "Point Forecast")
```

```{r}
plot <- ggplot(test, aes(month)) +
  geom_line(aes(y = stlf_clean, color = "Forecast")) +
  geom_line(aes(y = MonthlyRides, color = "Actual"))+
  scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) +
  ggtitle("STLF Monthly Rides 2024")
ggplotly(plot)
```


```{r}
arima_forecast <- stlf(ts_clean, h = 12, s.window = "periodic", method = "arima")
plot(arima_forecast)
```

```{r}
df_ar <- as.data.frame(arima_forecast)
test$ar_clean = df_ar$`Point Forecast`[1:10]
mape(test$MonthlyRides, test$ar_clean)
```

```{r}
df_ar <- df_ar %>% mutate(Month = row_number(),
                    ForCalc = Month -12,
                    Year = ifelse(ForCalc > 0, 2025, 2024),
                    Month = ifelse(ForCalc > 0, ForCalc, Month),
                    Category = "Forecast") %>%
  select(Year, Month, `Point Forecast`, `Lo 95`, `Hi 95`)
```

```{r}
#joined_data <- left_join(joined_data, df_ar, by = c("Year"= "Year", "Week"="Week"))  %>% rename("arima2" = "Point Forecast")
```

```{r}
plot <- ggplot(test, aes(month)) +
  geom_line(aes(y = ar_clean, color = "Forecast")) +
  geom_line(aes(y = MonthlyRides, color = "Actual"))+
  scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) +
  ggtitle("ARIMA Monthly Rides 2024")
ggplotly(plot)
```

**Hybrid Models**
```{r}
library(forecastHybrid)
mod1 <- hybridModel(ts_clean)
mod1_forecast <- forecast(mod1, h = 12)
plot(mod1_forecast)
```

```{r}
df_hybrid2 <- as.data.frame(mod1_forecast)
test$hybrid2 = df_hybrid2$`Point Forecast`[1:10]
mape(test$MonthlyRides, test$hybrid2)
```

```{r}
df_hybrid2 <- df_hybrid2 %>% mutate(Month = row_number(),
                    ForCalc = Month -12,
                    Year = ifelse(ForCalc > 0, 2025, 2024),
                    Month = ifelse(ForCalc > 0, ForCalc, Month),
                    Category = "Forecast") %>%
  select(Year, Month, `Point Forecast`, `Lo 95`, `Hi 95`)
```

```{r}
#joined_data <- left_join(joined_data, df_hybrid2, by = c("Year"= "Year", "Week"="Week"))  %>% rename("hybrid2" = "Point Forecast")
```

```{r}
plot <- ggplot(test, aes(month)) +
  geom_line(aes(y = hybrid2, color = "Forecast")) +
  geom_line(aes(y = MonthlyRides, color = "Actual"))+
  scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) +
  ggtitle("Hybrid2 Monthly Rides 2024")
ggplotly(plot)
```


**ARIMA/STLM**
```{r}
# ARIMA, STLM
mod_as_clean <- hybridModel(ts_clean, models = "as") 
as_fore_clean <- forecast(mod_as_clean, h = 12)
plot(as_fore_clean)
```

```{r}
df_as_clean <- as.data.frame(as_fore_clean)
test$as_clean = df_as$`Point Forecast`[1:10]
mape(test$MonthlyRides, test$as_clean)
```

```{r}
df_as_clean <- df_as_clean %>% mutate(Month = row_number(),
                    ForCalc = Month -12,
                    Year = ifelse(ForCalc > 0, 2025, 2024),
                    Month = ifelse(ForCalc > 0, ForCalc, Month),
                    Category = "Forecast") %>%
  select(Year, Month, `Point Forecast`, `Lo 95`, `Hi 95`)
```

```{r}
#joined_data <- left_join(joined_data, df_as_clean, by = c("Year"= "Year", "Week"="Week"))  %>% rename("as2_clean" = "Point Forecast")
```

```{r}
plot <- ggplot(test, aes(month)) +
  geom_line(aes(y = as2_clean, color = "Forecast")) +
  geom_line(aes(y = MonthlyRides, color = "Actual"))+
  scale_color_manual(values = c("Forecast" = "darkred", "Actual" = "darkblue")) +
  ggtitle("ARIMA + STLM Monthly Rides 2024")
ggplotly(plot)
```