---
title: "HomeworkSet_10"
author: "Joshua Tobey"
date: "2025-12-04"
output: html_document
---
```{r}
library(tidyverse)
library(tidymodels)
library(knitr)
library(ggplot2)

set.seed(445)
```
PROBELM 1) We will explore the maximal margin classifier on a toy data set.
a) 
Refer to the sketch in the pdf.

b) Equation for Optimal hyperplane, which I found in part d.  See the sketch in the pdf.

$f(x^*) =  \beta_0 + \beta_1x_1 + \beta_2x_2$
$>>1 - 2\beta_1 + 2\beta_2 = 0$

c) Describe the classification rule for the maximal marginal classifier. It should be along the lines of “Classify to Red if 𝛽0+𝛽1𝑋1+𝛽2𝑋2>0, and classify as Blue otherwise. Provide the values of 𝛽0,𝛽1,𝛽2.

\[
\beta_0 + \beta_1x_1 + \beta_2x_2 
\begin{cases}
> 0 & \text{Classify Red} \\
< 0 & \text{Classify Blue} \\
= 0 & \text{On hyperplane}
\end{cases}
\]

Values of $\beta_0, \beta_1 & \beta_2$: Refer to the pdf of sketch which has the algebra work on it.
It uses the 3 support vectors (2,1), (2,2) and (4,3) as well as their equations.

$\beta_0 = 1$
$\beta_1 = -2$
$\beta_2 = 2$

d) On your sketch, indicate the margin for the maximal margin classifier.

Refer back to pdf of sketch and algebra work.  Margins are in orange on the sketched graph.

Given hyperplane $ = 1 - 2\beta_1 + 2\beta_2 = 0$.  The margins are subject to:
\[
$1 - 2\beta_1 + 2\beta_2 = 1$
$1 - 2\beta_1 + 2\beta_2 = -1$
\]

e) Indicate the support vectors for the maximal margin classifier.

Testing which points give an equation equal to 1 for red points and -1 for blue points.
Red support vectors:
$(2,2) = 1 - 2(2) + 2(2) = 1$
$(4,4) = 1 - 2(4) + 2(4) = 1$
Blue support vectors:
$(2,1) = 1 - 2(2) + 2(1) = -1$
$(4,3) = 1 - 2(4) + 2(3) = -1$

f) Argue that a slight movement of the seventh observation would not affect the maximal margin hyperplane.

Seeing that the seventh is a blue categorized point, it lies far below the hyper plane, so slight adjustments won't cause the point to interfere with the support vectors that are much closer to the hyperplane.

g) Draw an additional observation on the plot so that the two classes are no longer separable by a hyperplane.

Refer to the pdf of the sketched graph.  The added point will be circled in green.

PROBLEM 2) We have seen that we can fit an SVM with a non-linear kernel in order to perform classification using a non-linear decision boundary. We will now see that we can also obtain a non-linear decision boundary by performing logistic regression using non-linear transformations of the features.

a) Generate a data set with 𝑛=500 and 𝑝=2, such that the observations belong to two classes with a quadratic decision boundary.

```{r}
n <- 500
x1 <- runif(n) - 0.5
x2 <- runif(n) - 0.5
y = as.numeric(x1^2 - x2^2 > 0)

```
b) Plot the observations, colored according to their class labels.
```{r}
plot(x1, x2, col = ifelse(y == 1, "red", "blue"))

```
c) Fit a logistic regression model to the data using 𝑋1 and 𝑋2 as predictors.
```{r}
library(parsnip)
log_df <- data.frame(x1 = x1, x2 = x2, response = factor(y))
log_spec <- logistic_reg()
log_fit <- log_spec |>
  fit(response ~ x1 + x2, data = log_df)
log_fit |>
  pluck("fit") |>
  summary()


```
I googled the error: 

Error in `check_outcome()`:
! For a classification model, the outcome should be a <factor>, not a double vector.
Run `rlang::last_trace()` to see where the error occurred.

This led me to using the factor() function on the y in creating a dataframe to fit the logistic model.

d) Apply this model to the training data in order to obtain a predicted class label for each training observation. Plot the observations, colored according to the predicted class labels. What shape is the decision boundary?

```{r}


```






